---
title: "Identifying promoter sequence architectures via a chunking-based algorithm using non-negative matrix factorisation"
author:
  - Sarvesh Nikumbh:
      email: s.nikumbh@imperial.ac.uk
      institute: [mrclms, icl]
      correspondence: true
  - Boris Lenhard:
      email: b.lenhard.imperial.ac.uk
      institute: [mrclms, icl]
      correspondence: true
institute:
  - mrclms: MRC London Institute of Medical Sciences, London, UK
  - icl: Institute of Clinical Sciences, Faculty of Medicine, Imperial College London, Hammersmith Hospital Campus, London, UK
geometry: "left=2cm,right=2cm,top=1.5cm,bottom=2cm"
bibliography: seqArchR-bibliography.bib
link-citations: yes
linkcolor: blue
output:
  bookdown::pdf_document2:
    includes:
      in_header: mystyles.sty
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: false
    anchor_sections: true
    number_tables: true
    keep_tex: true
    keep_md: true
    pandoc_args:
      - '--lua-filter=lua_filters/scholarly-metadata.lua'
      - '--lua-filter=lua_filters/author-info-blocks.lua'
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: console
---

<!-- Author affiliations handled using 
https://stackoverflow.com/a/52919256 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = 'svg', include=TRUE, eval=TRUE)
 
```



# Abstract {.unnumbered}

Core promoters are stretches of DNA at the beginning of genes that contain 
information that facilitates the binding of transcription initiation complex. 
Different functional subsets of genes have core promoters with distinct 
architectures and characteristic motifs. 
Some of these motifs inform the selection of transcription start sites (TSS). 
By discovering motifs with fixed distances from known TSS positions, we could 
in principle classify promoters into different functional groups.

Due to the variability and overlap of architectures, promoter classification 
is a difficult task that requires new approaches. 
In this study, we present a new procedure based on nonnegative matrix 
factorisation (NMF) and the associated software tool called seqArchR that 
cluster promoter sequences based on their motifs at near-fixed distances from 
a reference point, such as TSS. 
When combined with experimental data from CAGE, seqArchR can efficiently 
identify TSS-directing motifs, including known ones like TATA, DPE, and 
nucleosome positioning signal, as well as novel lineage-specific motifs and 
the function of genes associated with them. 
By using seqArchR on developmental time courses, we reveal how relative use 
of promoter architectures changes over time with stage-specific expression.

seqArchR is a powerful tool for initial genome-wide classification and 
functional characterization of promoters. 
Its use cases are more general: it can also be used to discover any motifs 
at near-fixed distances from a reference point, even if they are present in 
only a small subset of sequences. seqArchR is available at [http://www.bioconductor.org/packages/seqArchR](http://www.bioconductor.org/packages/seqArchR).


```{r set-paths-libs, include=FALSE}

library(reshape2)
library(patchwork)
library(readr)
if(requireNamespace("magick", quietly = TRUE)){
  library(magick)
}

library(ggplot2)
library(ggpubr)
library(motifStack)

## Note: always keep pdf to the end
use_device <- c("png", "tiff", "svg", "pdf")
use_tags <- 'A' # change this when different tag levels are required across all plots
use_nest <- '1' # for nested tags
fig_use_cores <- 16

figures_path <- file.path("figures")
data_path <- file.path("../experiments", "data")
results_path <- file.path("../experiments", "results")

archR_simdata_data_path <- file.path(data_path, "simulated-data")


archR_simdata_results_path <- file.path(results_path,
                                "simulated-data-archR-v0.1.8-bound-1e08", 
                                paste0(fig_use_cores, "cores"))

archR_simdata_results_path_mem <- file.path(results_path,
                                "simulated-data-archR-v0.1.8-bound-1e08", 
                                "48cores")


nplb_simdata_results_path <- file.path(results_path, "comparison-approaches",
                                paste0(fig_use_cores, "cores"), "simulated-data")

### Drosophila melanogaster
### 
dm_sample_names <- paste0("RAL28_", c("2_to_4", "6_to_8", "10_to_12_sample12"))
archR_dm_results_path <- file.path(results_path,
                                "drosophila-schor2017", "with_archR_v0.1.8")

archR_dm_chen_results_path <- file.path(results_path,
                                "drosophila-chen2014", "with_archR_v0.1.8")

dm_result_serial_run_path <- file.path(archR_dm_chen_results_path, 
    "archR_result_drosophila-chen2014_serial_run_modSelType_stability_chunkSize_2000_bound_1e-08_collate_FTFFF")

## store result_directory path
dm_result_dir_path <- vector("list", length(dm_sample_names)) 
names(dm_result_dir_path) <- dm_sample_names

for(sn in dm_sample_names){
    dm_result_dir_path[[sn]] <- file.path(archR_dm_results_path, 
                                       paste0(sn, "_results"))
}


### Danio rerio
### 
zf_sample_names <- c("64_cells", "dome_30perc_epiboly", "prim6")
archR_zf_results_path <- file.path(results_path,
                                "zebrafish-nepal2013", "with_archR_v0.1.8")

## store result_directory path
zf_result_dir_path <- vector("list", length(zf_sample_names)) 
names(zf_result_dir_path) <- zf_sample_names

for(sn in zf_sample_names){
    zf_result_dir_path[[sn]] <- file.path(archR_zf_results_path, 
                                       paste0(sn, "_results"))
}


### Homo sapiens
### 
hg_sample_names <- c("human_cellGroup_merged")
archR_hg_results_path <- file.path(results_path, "human")

## store result_directory path
hg_result_dir_path <- vector("list", length(hg_sample_names)) 
names(hg_result_dir_path) <- hg_sample_names

for(sn in hg_sample_names){
    hg_result_dir_path[[sn]] <- file.path(archR_hg_results_path, 
                                       paste0(sn, "_results"))
}


```

# Background {#background}

Promoter sequences are stretches of DNA flanking the transcription start site 
(TSS) of a gene.
Typically, a window spanning $-40$ to $+40$ base pairs (bp) (upstream denoted 
by '$-$' and downstream by '$+$') relative to the TSS is called the core 
promoter sequence.
Further upstream from $-40$ to $-250$ bp is called the proximal promoter 
region [@butler2002rna]. 
These regions work together with enhancers \textemdash\space which can be 
anywhere in the genome, upstream or downstream of the gene they affect 
\textemdash\space to regulate gene expression. 
Transcription of genes by RNA polymerase II (RNA PolII) happens by assembling 
the pre-initiation complex of RNA PolII and different basal transcription 
factors (TFs) [@kadonaga2012perspectives].
These TFs recognise different sequence motifs present in the core promoters 
for binding.


However, there is no universal promoter architecture. 
Not all motif elements are present in all core promoters.
For instance, \texttt{TATA}-box, which is bound by the \texttt{TATA}-box binding 
protein (TBP), is a popular textbook example of a core promoter motif,
but only \~20% of the known core promoters of protein-coding genes in animals 
contain a TATA-box [@kadonaga2012perspectives]. 
Furthermore, transcription initiation can be 'focused' or 'dispersed' 
[@carninci2006genome; @juven2010regulation]. 
In focused transcription, there is usually a single nucleotide (nt) position 
where most of the transcription begins. 
In dispersed transcription there are multiple weaker start positions spread 
up to \~100 bp within the promoter, and the dominant position is determined 
by the position of $+1$ nucleosome [@haberle2014shifting; @dreos2016influence].
The resultant promoters are thus called *sharp* and *broad* respectively.
Sharp promoters are associated with tissue-specific genes, while 
broad promoters are associated with constitutively expressed genes and a subset 
of tissue restricted ones.
The most important realisation about different promoter architectures is that 
there is a TSS-selection determining sequence element at near-fixed distance 
from the dominant TSS (Figure \@ref(fig:archR-schematic-bio)).
Different TSS selection mechanisms suggest that there are multiple fundamentally 
different promoter architectures, and that they are functionally specialised.
Heterogeneity in architectures at the organism-level gives rise to additional 
diversity [@carninci2006genome;@narlikar2014multiple;@haberle2016promoter].
Also, the advent of chromatin interaction experiments in the last decade has 
enabled better exploration of the enhancer-promoter interaction specificity.
Thus, studying and defining the promoter sequence architectures that give rise 
to such specificities in transcription regulation mechanisms is an important step.

```{r archR-schematic-bio, fig.align="center", out.width = "100%", out.height="30%", echo=FALSE, include=TRUE, fig.cap="Schema depicting how, across organisms, different promoter architectures have a TSS-selection determining sequence element at near-fixed distance from their dominant transcription start sites. nt = Nucleotide(s)", warning=FALSE, message=FALSE}

cage_signal_df <- data.frame(x    = c(49, 50, 50, 51, 50, 51, 44, 46, 48, 50, 51, 53, 55, 56, 57, 60), 
                             xend = c(49, 50, 50, 51, 50, 51, 44, 46, 48, 50, 51, 53, 55, 56, 57, 60),
                             y    = c(c(40, 40), c(30, 30), c(20, 20), c(10, 10, 10, 10, 10, 10, 10, 10, 10, 10)), #bottoms
                             yend = c(c(41, 45), c(35, 32), c(25, 21), c(13, 12, 11, 15, 14, 13, 12, 13, 11, 11))) #tops

nucl_df <- data.frame(x0 = c(110, 141), 
                      y0 = rep(5, 2),
                      r = rep(4.5, 2),
                      a = rep(15, 2),
                      b = rep(4.5, 2))

## Generate TATA motif
foo <- Biostrings::readDNAStringSet(file.path(dm_result_dir_path[[3]], "TATA_subsequences.fa"))
foo_pcm <- Biostrings::consensusMatrix(foo)[1:4,]
tata_pcm <- new("pcm", mat=as.matrix(foo_pcm), name="TATA-box")

## Generate DPE motif
foo <- Biostrings::readDNAStringSet(file.path(dm_result_dir_path[[3]], "DPE_subsequences.fa"))
foo_pcm <- Biostrings::consensusMatrix(foo)[1:4,]
dpe_pcm <- new("pcm", mat=as.matrix(foo_pcm), name="DPE")

## Generate W-box motif
foo <- Biostrings::readDNAStringSet(file.path(zf_result_dir_path[[1]], "Wbox_subsequences.fa"))
foo_pcm <- Biostrings::consensusMatrix(foo)[1:4,]
wbox_pcm <- new("pcm", mat=as.matrix(foo_pcm), name="W-box")

## Generate periodicity signal
foo <- Biostrings::readDNAStringSet(file.path(zf_result_dir_path[[3]], "nuc_pos_periodicity_subsequences.fa"))
foo_pcm <- Biostrings::consensusMatrix(foo)[1:4,]
nuc_pcm <- new("pcm", mat=as.matrix(foo_pcm), name="Phasing by +1 nucleosome")


# motifs_df <- data.frame(xmin=c(15, 13, 75), ymin=c(40, 30, 20), xmax=c(28, 28, 87), ymax=c(48, 50, 28))

motifs_df <- data.frame(x=c(20, 20, 78, 120), y=c(43, 45, 24, 30), width=c(15, 15, 13, 55), height=c(6, 30, 8, 40))
motifs_df$motif <- list(motifStack::pcm2pfm(tata_pcm), 
                        motifStack::pcm2pfm(wbox_pcm),
                        motifStack::pcm2pfm(dpe_pcm),
                        motifStack::pcm2pfm(nuc_pcm))

##
signal_col <- RColorBrewer::brewer.pal(n = 5, name = "Blues")[3]
nucl_col <- RColorBrewer::brewer.pal(n = 5, name = "Accent")[1]

pl <- ggplot() + geom_blank() + xlim(-50, 100) + ylim(0, 50) + 
      theme_classic() +
      ggplot2::scale_x_continuous(labels = seq(-50, 100, by = 10), breaks = seq(0, 150, by = 10)) +
      ggplot2::annotate(geom = "segment", 
          x = 0, xend=150, y=seq(10,40, 10), yend=seq(10,40, 10), linewidth=0.5) +
      ## draw CAGE signal
      ggplot2::annotate(geom = "segment", color = signal_col,
          x = cage_signal_df$x, xend=cage_signal_df$xend, 
          y = cage_signal_df$y, yend=cage_signal_df$yend, linewidth=1.5) + 
      ## show dTSS
      ggplot2::annotate(geom = "text", label = "*", x = 50, y = -1+c(39, 29, 19, 09), size = 7) +
      ## text legend of dTSS
      ggplot2::annotate(geom = "text", label = "*", x = 124, y = c(46.5), size = 7) +
      ggplot2::annotate(geom = "text", label = "dominant TSS", x = 135, y = c(47)) + 
      ## add nucleosomes
      # ggforce::geom_circle(aes(x0 = x0, y0 = y0, r = r, fill = r, alpha = 0.5), data = nucl_df) +
      ggforce::geom_ellipse(aes(x0 = x0, y0 = y0, a = a, b = b, 
                            fill = r, alpha = 0.5, angle = 0), data = nucl_df) +
      ggplot2::scale_fill_distiller(palette = "Accent") +
      ggplot2::annotate(geom = "segment", 
          x = c(95, 110, 126, 141), xend = c(125, 110, 156, 141), 
          y = c(5, 0.5, 5, 0.5), yend = c(5, 9.5, 5, 9.5), alpha = 0.2) +
      ## 
      ggplot2::coord_fixed() +
      ## TATA
      # motifStack::geom_motif(data = motifs_df, use.xy = FALSE,
      #                       mapping = aes(xmin=xmin, ymin=ymin, xmax=xmax, ymax=ymax, 
      #                       motif=motif, fontfamily = "Arial")) +
      motifStack::geom_motif(data = motifs_df, use.xy = TRUE,
                            mapping = aes(x = x, y = y, width = width, height = height,
                            motif=motif, fontfamily = "Arial")) +
      ## text legend nucleosome
      ggplot2::annotate(geom = "text", label = "Nucleosome", x = 86, y = 1) + 
      ggplot2::annotate(geom = "segment", linetype = "dashed", x = 50, xend = 95.5, y = 8, yend = 8) + 
      ## other dashed lines
      ggplot2::annotate(geom = "segment", linetype = "dashed", x = 50, xend = 18, y = 38, yend = 38) + 
      ggplot2::annotate(geom = "segment", linetype = "dashed", x = 50, xend = 18, y = 28, yend = 28) + 
      ggplot2::annotate(geom = "segment", linetype = "dashed", x = 50, xend = 75, y = 18, yend = 18) + 
      ## TATA, W-box and DPE
      ggplot2::annotate(geom = "text", label = "TATA-box", x = 15, y = 50) + 
      ggplot2::annotate(geom = "text", label = "30-33 nt", x = 32, y = 36) + 
      ggplot2::annotate(geom = "text", label = "W-box", x = 10, y = 33) + 
      ggplot2::annotate(geom = "text", label = "30-33 nt", x = 32, y = 26) +
      ggplot2::annotate(geom = "text", label = "DPE", x = 70, y = 27) + 
      ggplot2::annotate(geom = "text", label = "28 nt", x = 62, y = 16) +
      ggplot2::annotate(geom = "text", label = "50 nt", x = 76, y = 6) +
      ##
      ggplot2::theme(legend.position="none") + 
      ggeasy::easy_remove_axes(which = c("both"),
                               what = c("ticks", "title", "text", "line"))
##

use_fname <- file.path(figures_path, "seqArchR-figure1-bio-schematic.pdf")
ggplot2::ggsave(filename = use_fname, plot = pl, device = "pdf", 
                width = 9, height = 3.5)

if(file.exists(use_fname) && !knitr::is_html_output()){
    knitr::include_graphics(use_fname)
}
```

Promoter sequence architectures are not always characterised just by occurrences 
of short sequence motifs but at times also by the sequence composition, for 
example, \texttt{GC}-richness or CpG island promoters.
Using motif finding algorithms to process promoter sequences, as is traditional, 
has a limitation.
First, they cannot identify differences in architectures stemming from sequence 
compositions.
Second, in scenarios where promoter sequences are indeed characterised by motifs, 
scanning the sequences for motifs from a database (when available) limits the 
study to looking at only known motifs.
These known motifs often do not explain much of the diversity among the promoter 
sequences.
In case of novel organisms for whom no databases of known motifs exist yet, a 
*de novo* motif search can be performed.
But most motif finding algorithms expect the motifs to be statistically 
over-represented in the promoter sequences in comparison to background sequences.
Such approaches can fail to identify motifs that are present in a relatively 
small subset of sequences and therefore not statistically `significant'.
They do not provide precise position information of the motifs.
Furthermore, they also often miss discovering low-complexity or degenerate 
motifs.
Thus, approaches that can overcome the above limitations in using motif finding 
algorithms and identify architectures in promoter sequences *de novo* are 
required to study promoter sequence architectures.

Here, we present seqArchR, an unsupervised approach using non-negative matrix 
factorisation (NMF) [@lee1999learning] for *de novo* characterisation of 
diversity in promoter sequence architectures. 
seqArchR does not require background sequences.
It needs no information on the kind (motifs- or sequence composition-based) or 
number of architectures expected to be present in the input sequences.
When architectures characterised by motifs may be present, it does not require 
any specifications on the number of motifs, their sizes or any gaps in motifs 
or between them when they work in a cooperative fashion.
seqArchR's chunking-based approach can unravel architectures harboured by 
very few sequences out of the whole input.
This is often helpful to precisely tease out minor positional variations in 
features which can be functionally relevant.
For instance, the spacing between the `TATA`-box and the TSS is known to 
correlate with tissue specificity [@ponjavic2006transcriptional].

```{r serial-run-chen2014, include=FALSE, echo=FALSE}

chen2014_result_serial <- readRDS(file.path(dm_result_serial_run_path, 
    "archRresult.rds"))

unlist(chen2014_result_serial$timeInfo)
cumsum(chen2014_result_serial$timeInfo)

```

seqArchR is inspired by 'No promoter left behind' (NPLB) [@mitra2016nplb], 
the current state-of-the-art method identifying such sequence architectures 
*de novo* from aligned promoter sequences. 
NPLB uses a Bayesian approach to jointly learn features and important positions 
in an architecture. 
It identifies the number of architectures best explaining the complete set of
promoter sequences by  cross-validation where the model with the highest 
loglikelihood is chosen.
NPLB has a relatively long execution time making it less effective as an 
exploratory tool.
It takes about 10 hours to process ca. 6600 promoter sequences in Drosophila
compared to seqArchR’s 42 minutes when run serially to obtain comparable results.
Additionally, seqArchR gives the user complete control over the granularity of 
identified clusters which is often useful in the exploratory setting.



Since @lee1999learning's work, NMF has become a popular technique finding wide 
applications across domains. 
Some earlier applications of NMF in biology include those for inference of 
biological processes [@brunet2004metagenes], sequence motifs characterisation 
[@previousNMFonSeqs], and multi-omics analysis [@stein2018enter].
In particular, @previousNMFonSeqs use NMF for characterizing sequence data 
sets by identifying position-dependent motifs in them.
In their approach, the input matrix encodes the counts of the *k*-mers and 
their occurrence positions in the given collection of sequences.
This approach assumes that all sequences in the data set contain the same 
motifs or minor variations.
Although this approach suits well for identifying protein-DNA binding sites 
from ChIP-seq like data, it assumes a single mode of binding and does not cater 
to the scenario of multiple modes of protein-DNA binding [@mumodLeelavati].
In comparison, seqArchR makes no such assumptions and models the collection 
of sequences to harbour any number of architectures, each with an unknown 
number of motifs\textendash\space or non-motifs\textendash based sequence 
features. 

This paper describes seqArchR in complete detail and presents results from 
computational experiments demonstrating its efficacy.
We compare seqArchR with NPLB in analyzing a simulated dataset where the 
ground truth of clusters/architectures is known, and a set of CAGE-based core 
promoter sequences from @chen2014comparative (Supplementary). 
Furthermore, we show results from processing CAGE-based promoter sequences in 
fruit fly [@schor2017promoter], zebrafish [@nepal2013dynamic] and human 
[@encode2012integrated; @encodedataportal] to demonstrate seqArchR's ability 
to seamlessly identify heterogeneous architectures across organisms with large 
differences in promoter sequence composition.
seqArchR is provided as an R/Bioconductor package accessible at 
[https://www.bioconductor.org/packages/seqArchR](https://www.bioconductor.org/packages/seqArchR) 
or [https://snikumbh.github.io/seqArchR](https://snikumbh.github.io/seqArchR).




# Results {#results}






## seqArchR: a novel algorithm using non-negative matrix factorisation for *de novo* identification of sequence architectures

```{r archR-algo-figure, fig.align="center", out.width="100%", out.height="40%", echo=FALSE, include=TRUE, fig.cap="Overview of seqArchR. (\\textbf{A}) Schematic showing seqArchR's chunking-based, iterative algorithm. (\\textbf{B}) Schematic describing input to seqArchR and the factorisation output. For each chunk of sequences being processed with NMF, the sequences are represented as a one-hot encoded matrix (hence, $0/1$ matrix), denoted in the schematic by matrix $V_{p\\times n}$; matrices $W_{p\\times k}$ and $H_{k \\times n}$ are respectively the basis matrix and coefficients matrix obtained upon factorisation. $n$ denotes the number of input sequences, $p$, the number of features, $k$, the optimal number of dimensions selected for the low-rank representation and $L$, the length of the input sequences. The schematic depicts one-hot encoding of the dinucleotide profile of sequences. One can use the mono- or dinucleotide profile of sequences."}

###
# png_fname <- file.path("figures", "archR_algorithm_static_illustrator_cropped.png")
# if(file.exists(png_fname) && knitr::is_html_output()){
#     knitr::include_graphics(png_fname)
# }

### 
pdf_fnames <- paste0(file.path(figures_path, 
                    c("archR_algorithm_landscape_no_box-crop.pdf", 
                        "archR_figure1_part_no_box-crop.pdf")))
# fig1_pdf_name <- file.path(figures_path, "samarth_archR_figure1.pdf")
use_fname <- file.path(figures_path, "archR_figure1_no_box_new-crop.pdf")

if(FALSE && requireNamespace("magick", quietly = TRUE)){
  if(requireNamespace("pdftools", quietly = TRUE)){
    # pdf_fnames <- paste0(file.path(figures_path, 
    #                               c("archR_algorithm_landscape_no_box-crop.pdf",
    #                                 "archR_figure1_part_no_box-crop.pdf")))

    
    ## algorithm figure part
    fig1 <- cowplot::ggdraw() + 
            cowplot::draw_image(magick::image_read_pdf(pdf_fnames[1], density = 600))
    
    ## NMF explanation figure part
    fig2 <- cowplot::ggdraw() + 
            cowplot::draw_image(magick::image_read_pdf(pdf_fnames[2], density = 600))
    
    comb_fig <- cowplot::plot_grid(fig1, NULL, fig2, 
                                   label_fontface = "plain", label_size = 8, 
                                   labels = c("A", NULL, "B"), 
                                   rel_widths = c(1, 0.1, 1),
                                   align="hv", nrow = 1, 
                                   vjust = c(4,4), hjust = c(0, -2))
    
    ## May have to separately crop the figure again using pdfcrop on the 
    ## command line                       
    
    cowplot::ggsave2(use_fname, 
                     plot = comb_fig, device = "pdf", dpi = 600, 
                     width = 15, height = 15, units = "cm")
    # comb_fig
  }
}

if(knitr::is_html_output()){
    comb_fig
}

if(file.exists(use_fname) && !knitr::is_html_output()){
    # use_fig1_pdf_name <- file.path(figures_path, "archR_figure1_no_box_new-crop.pdf")
    knitr::include_graphics(use_fname)
}




```

To classify promoters based on their fixed-position determinants, we developed, 
seqArchR , a chunking-based iterative algorithm using NMF for *de novo* 
identification of architectural elements.
The input to seqArchR is a $(0,1)$-matrix which is a one-hot encoded 
representation of dinucleotide profiles of a gapless alignment of DNA sequences.
As shown in the schematic in Figure \@ref(fig:archR-algo-figure), seqArchR 
processes the whole collection of input sequences one chunk (subset of sequences) 
at a time.
The $(0,1)$-matrix for each chunk of sequences is processed with NMF.
NMF decomposes the matrix into two low-rank matrices \textemdash\space the 
basis matrix and the coefficients matrix.
Columns of the basis matrix (a.k.a. basis vectors) represent the different 
potential architectures, and along its rows are the loadings for the features 
per architecture (see Figure \@ref(fig:archR-algo-figure), panel B).
On the other hand, in the coefficients matrix, where each column corresponds 
to one sequence, its rows hold the per-sequence coefficients for each 
architecture.
A sequence with a high coefficient for an architecture is expected 
to harbour features captured by that architecture.
For example, consider matrix $H_{k\times n}$ in Figure 
\@ref(fig:archR-algo-figure)B, the first and second sequence have the highest 
coefficient for architecture I among all four architectures, and are expected 
to harbour sequence features prominent for architecture I (see corresponding 
to the sequence logos on the top). 
Because the input matrix already encodes the position information of the 
nucleotides in the sequences, the factorisation identifies the characteristic 
architectures (combinations of nucleotides over-represented at specific 
positions) describing different subset of sequences.

Per chunk, seqArchR performs a model selection procedure to find the 
appropriate number of basis vectors suitable to represent the set of sequences 
in the chunk in a lower-dimensional space.
Two model selection procedures are available: stability\textendash\space and 
cross-validation\textendash based (see Section [Methods](#methods) for details).
This decomposition is used to obtain clusters of sequences characterised by a 
diverse set of sequence features.


After all chunks have been factorised and sequence clusters obtained, similar 
clusters from different chunks are collated.
Similarity of clusters is judged based on the similarity of corresponding NMF 
basis vectors
(see [Methods](#methods) section for additional details). 
The collated set of clusters may need further processing to weed out any 
mis-assignments or to crisply identify architectures with shared motifs or 
those with minor positional shifts.
In this case, seqArchR iterates over the collated clusters treating each of 
them as separate chunks and repeating the above steps.
The number of iterations required depends on the structure in the input data.
The algorithm is described in detail in the [Methods](#methods) section.



## seqArchR is fast and attains high accuracy on simulated data {#sim-data-results}

| Cluster |               Motifs                | Occurrence position | \#Sequences |
|---------|:-----------------------------------:|:-------------------:|:-----------:|
| A       |   Dinucleotide repeat \texttt{AT}   |     every 10 nt     |     200     |
| B       | \texttt{GATTACA} and \texttt{GAGAG} |      40 and 60      |     350     |
| C       |           \texttt{GAGAG}            |         60          |     150     |
| D       |  \texttt{TCAT} and \texttt{GAGAG}   |      40 and 80      |     300     |

: (\#tab:table-simulated-data) Sequence motifs corresponding to four architectures/clusters in the simulated data.

In order to demonstrate seqArchR's efficacy, we performed computational 
experiments on simulated data as follows.
We generated a set of 1000 simulated DNA sequences, each 100 nucleotides long.
All nucleotides at any position in these sequences appeared with a uniform 
probability of 0.25.
We planted different motifs at specific positions in these sequences such that 
they resulted in four clusters with as many architectures.
These are described in Table \@ref(tab:table-simulated-data).



All motifs except those in cluster A were planted with mutations, as follows.
For planting a motif in a sequence, any one position in the motif was randomly 
selected for mutation. 
The nucleotide at this position was then mutated with probability (mutation 
rate) 
{`r mutpos <- 1:3;mutrate <- seq(0.1,0.5,by=0.1);paste(mutrate, collapse=", ")`}.
This resulted in five variations of DNA sequences -- one mutated position 
$\times$ five mutation rates for all motifs. 
This procedure was repeated, selecting two and three positions for mutation 
per motif (instead of one), giving us a total of 15 variations.
For each of these variations, we performed experiments with chunk sizes 
{`r chunkSizes <- c(200,500,1000);paste(chunkSizes, collapse = ", ")`}.
Thus, all parameter combinations resulted in a total of 
`r length(mutpos)*length(mutrate)*length(chunkSizes)` datasets to be processed 
by seqArchR.
For robustness, each of these was repeated 10 times with different random seeds.
Here, we used stability-based model selection.


For comparison, we also processed these `r length(mutpos)*length(mutrate)` 
datasets with NPLB [@mitra2016nplb].
Performance results of seqArchR and NPLB are presented next.
All reported timing information is for experiments performed in parallel on 
`r fig_use_cores` cores of Intel(R) Xeon(R) Gold 6154 CPUs.

```{r read-input-TSVfile, include=FALSE}

filename <- file.path(archR_simdata_results_path, "archR_result_simulated_data_summary.tsv")
archR_simdata_results <- readr::read_tsv(file = filename, col_names = TRUE, 
                                        col_types = cols())

filenames <- list.files(path = archR_simdata_results_path_mem,
    pattern = "*_mem.*.tsv", full.names = TRUE)

mem_tsv_list <- lapply(filenames, function(f){
    temp_tsv <- readr::read_tsv(file = f, col_names = TRUE, col_types = cols())
    temp_tsv
})

archR_simdata_results_mem <- dplyr::bind_rows(mem_tsv_list)
archR_simdata_results_mem$index <- seq(nrow(archR_simdata_results_mem))


######
filename <- file.path(nplb_simdata_results_path, "nplb_result_simulated_data_summary.tsv")
nplb_simdata_results <- readr::read_tsv(file = filename, col_names = TRUE, 
                                        col_types = cols())

# Rename columns? because ggpubr doesn't have provision for renaming them in 
# the legend on the fly

# head(archR_simdata_results)
```



**Performance as adjusted Rand index (ARI).** 
Performance is reported using the adjusted Rand index (ARI) which measures the 
similarity between two clustering solutions or a clustering and ground truth 
(when available), and is corrected for chance.
As seen in Figure \@ref(fig:simdata-results)A, irrespective of the chunk size,
seqArchR attains a high ARI value when the mutation rate and the number of 
mutated motif positions is low.
For any given chunk size, the performance degrades with increase in the mutation
rate, while for the same mutation rate, increasing the number of mutated motif 
positions hardly affects performance.
For smaller chunk size, seqArchR's performance shows a slight increase in 
variance with increase in mutation rate.
With a larger chunk size, seqArchR demonstrates very low variance in 
comparison to NPLB which shows high variance for all mutation rates.

Figures \@ref(fig:simdata-results)E and F show the time taken (in minutes) 
by seqArchR and NPLB.
In Figure \@ref(fig:simdata-results)E, the colours denote time taken by 
seqArchR to complete different iterations.
We observe that processing the whole data in many, small chunks takes more 
time overall.
This is expected: for each chunk, the stability-based model selection procedure 
starts from checking for just two clusters and continues checking further values 
until the bounding threshold on the instability of identified cluster features 
is satisfied (See more details of the algorithm in the [Methods](#archR-algorithm) 
section).
Since sequences from each cluster are distributed uniformly randomly across 
the complete set of sequences, for each chunk, the optimum number of clusters 
is expected to be around the true number of clusters.
Thus, there is some redundancy of checking the lower values for every chunk.
This redundancy reduces for larger chunk sizes.


```{r ggpubr-ari-facet-grid, warning=FALSE}

relevant_idx <- dplyr::filter(archR_simdata_results, nSeqs == 1000)$index

only_final <- TRUE ## set FALSE if ARI values for each iteration should be plotted

ari_cols <- c('ARI_Iter1', 'ARI_Iter2', 'ARI')

    
ari_melted_df <- reshape2::melt(archR_simdata_results[relevant_idx,], 
                    id.vars = c('chunkSize', 'mutpos', 'mutrate'),
                    measure.vars = ari_cols[ifelse(only_final, 3, seq_len(3))]
                )
ari_melted_df$chunkSize <- ari_melted_df$chunkSize/1000

paired_blues <- RColorBrewer::brewer.pal(3, "Paired")[1:2]

archR_ari_pl1 <- ggpubr::ggbarplot(data = ari_melted_df, x = "mutrate", 
                    y = "value",
                    fill = paired_blues[1],
                    size = 0.5, 
                    color = paired_blues[1],
                    facet.by = c("mutpos", "chunkSize"), 
                    legend = ifelse(only_final, "top", "top"),
                    ylim = c(0.25,1.05),
                    xlab = "Mutation rate", 
                    ylab = "Adjusted Rand index (ARI)",
                    ggtheme = ggpubr::theme_pubr(),#theme_bw(base_size = 12),
                    legend.title = "Stage",
                    ## add params
                    add = c("mean_sd", "jitter"),
                    add.params = list(size = 0.8, color = paired_blues[2]),
                    ## label params
                    label = TRUE, lab.pos = "out", lab.vjust = -1.5,
                    lab.nb.digits = 2, lab.size = 3.5,
                    ##
                    panel.labs = list(
                        chunkSize = paste0("Chunk size = ", 
                            levels(as.factor(ari_melted_df$chunkSize)), "K"),  
                        mutpos = paste("p =", 1:3)),
                    x.text.angle = 45, palette = "Paired")

nSeqs_val <- c(5000, 10000)



relevant_idx <- dplyr::filter(archR_simdata_results, nSeqs == c(5000, 10000))$index
ari_melted_df2 <- reshape2::melt(archR_simdata_results[relevant_idx,],
            id.vars = c('chunkSize', 'nSeqs'),
            measure.vars = ari_cols[ifelse(only_final, 3, seq_len(3))]
            )
ari_melted_df2$nSeqsInK <- ari_melted_df2$nSeqs/1000
ari_melted_df2$chunkSizeInK <- ari_melted_df2$chunkSize/1000

archR_ari_pl2 <- ggpubr::ggbarplot(data = ari_melted_df2, 
                x = "chunkSizeInK", y = "value",
                fill = paired_blues[1], 
                size = 0.5, 
                color = paired_blues[1],
                facet.by = c("nSeqsInK"),
                legend = "none",
                ylim = c(0.25,1.1), 
                xlab = paste0("Chunk size ", expression('(x 1K)')),
                ylab = "ARI",
                ggtheme = ggpubr::theme_pubr(),#theme_bw(base_size = 12),
                legend.title = "Stage",
                ## add params
                add = c("mean", "jitter"),
                add.params = list(size = 0.8, color = paired_blues[2]),
                ## label params
                label = TRUE, lab.pos = "out", lab.vjust = -1.5,
                lab.nb.digits = 2, lab.size = 3.5,
                ##
                panel.labs = list(nSeqsInK = 
                                    paste0("n = ",
                                           as.numeric(levels(as.factor(
                                             ari_melted_df2$nSeqsInK
                                             ))),
                                           "K")),
                x.text.angle = 0, palette = "Paired",
                scales = "free_x", nrow = 2, ncol = 1
                )



####### ari plot for NPLB

relevant_idx <- dplyr::filter(nplb_simdata_results, nSeqs == 1000 & nClusters != -1)$index

ari_melted_df <- reshape2::melt(nplb_simdata_results[relevant_idx,], 
                    id.vars = c('nSeqs', 'mutpos', 'mutrate'),
                    measure.vars = ari_cols[3]
                )
ari_melted_df$nSeqs <- ari_melted_df$nSeqs/1000

paired_col <- RColorBrewer::brewer.pal(8, "Paired")[7:8]

nplb_ari_pl1 <- ggpubr::ggbarplot(data = ari_melted_df, x = "mutrate", y = "value",
                    fill = paired_col[1],
                    size = 0.5, 
                    color = paired_col[1],
                    facet.by = c("mutpos", "nSeqs"), 
                    legend = ifelse(only_final, "top", "top"),
                    ylim = c(0.25,1.05),
                    xlab = "Mutation rate", 
                    ylab = FALSE,#  "Adjusted Rand index (ARI)",
                    ggtheme = ggpubr::theme_pubr(),#theme_bw(base_size = 12),
                    legend.title = "Stage",
                    ## add params
                    add = c("mean_sd", "jitter"),
                    add.params = list(size = 0.8, color = paired_col[2]),
                    ## label params
                    label = TRUE, lab.pos = "out", lab.vjust = -1.5,
                    lab.nb.digits = 2, lab.size = 3.5,
                    ##
                    panel.labs = list(
                        nSeqs = paste0("n = 1K"),   
                        mutpos = paste("p =", 1:3)),
                    x.text.angle = 45,
                    palette = "Paired")


##
relevant_idx <- dplyr::filter(nplb_simdata_results, nSeqs != 1000 & nClusters != -1)$index
ari_melted_df2 <- reshape2::melt(nplb_simdata_results[relevant_idx,],
            id.vars = c('nSeqs'),
            measure.vars = ari_cols[3]
            )
ari_melted_df2$nSeqs <- ari_melted_df2$nSeqs/1000
##
nplb_ari_pl2 <- ggpubr::ggbarplot(data = ari_melted_df2, 
                x = "nSeqs", y = "value", width = 0.4,
                fill = paired_col[1], 
                size = 0.5, 
                color = paired_col[1],
                legend = "none",
                ylim = c(0.25,1.05), 
                xlab = paste0("#Sequences ", expression('(x 1K)')),
                ylab = "ARI",
                ggtheme = ggpubr::theme_pubr(),
                legend.title = "Stage",
                ## add params
                add = c("mean_sd", "jitter"),
                add.params = list(size = 0.8, color = paired_col[2]),
                ## label params
                label = TRUE, lab.pos = "out", lab.vjust = -1.5,
                lab.nb.digits = 2, lab.size = 3.5,
                ##
                x.text.angle = 0, palette = "Paired"
                )
    

####### Time info plot
copy_simdata_results <- archR_simdata_results
colnames(copy_simdata_results)[9:10] <- paste0(1:2)#
#
nSeqs_val <- c(1000, 5000, 10000)
paired_blues <- RColorBrewer::brewer.pal(3, "Paired")[2:1]

relevant_idx <- dplyr::filter(copy_simdata_results, nSeqs > 100)$index

time_melted_df2 <- reshape2::melt(copy_simdata_results[relevant_idx,],
                id.vars = c('chunkSize', 'nSeqs'),
                measure.vars = paste0(1:2)
                )
time_melted_df2$chunkSizeInK <- time_melted_df2$chunkSize/1000
time_melted_df2$nSeqsInK <- time_melted_df2$nSeqs/1000
time_melted_df2$value <- time_melted_df2$value/60
## ^ All timings are in seconds, convert to minutes



archR_time_pl <- ggpubr::ggbarplot(data = time_melted_df2,
                x = "chunkSizeInK", y = "value",
                fill = "variable", 
                width = 0.4, color = "variable",
                facet.by = c("nSeqsInK"), 
                ncol = length(levels(as.factor(
                                             time_melted_df2$nSeqsInK
                                             ))),
                position = ggplot2::position_stack(reverse = TRUE),
                ## add params
                add = "mean",
                ## label params, if any
                ##
                xlab = paste0("Chunk size ", expression('(x 1K)')),
                ylab = "Time (minutes)", 
                legend.title = "seqArchR Iteration",
                legend = c(0.18, 0.9),
                font.legend = c(10),
                panel.labs = list(nSeqsInK = 
                                    paste0("#Sequences = ",
                                           as.numeric(levels(as.factor(
                                             time_melted_df2$nSeqsInK
                                             ))),
                                           "K")),
                ggtheme = ggpubr::theme_pubr(),
                x.text.angle = 0, palette = paired_blues,
                scales = "free_x"
                )



archR_time_pl <- archR_time_pl  + 
                  scale_y_continuous(sec.axis = ggplot2::dup_axis(name = NULL)) +
                  ggplot2::theme(legend.direction = "horizontal")


####### NPLB time plots
##
relevant_idx <- dplyr::filter(nplb_simdata_results, nClusters != -1)$index
time_melted_df2 <- reshape2::melt(nplb_simdata_results[relevant_idx,],
            id.vars = c('nSeqs'),
            measure.vars = "runTime_m"
            )
time_melted_df2$nSeqs <- time_melted_df2$nSeqs/1000
##
paired_col <- RColorBrewer::brewer.pal(8, "Paired")[7:8]
nplb_time_pl <- ggpubr::ggbarplot(data = time_melted_df2, 
                x = "nSeqs", y = "value", 
                fill = paired_col[1], width = 0.4, 
                color = paired_col[1], # sets border color of bars
                ## add params
                add = c("mean"), 
                add.params = list(size= 0.3, color = paired_col[2]),
                ## label params, if any
                xlab = paste0("#Sequences ", expression('(x 1K)')),
                ylab = "Time (minutes)",
                legend.title = "",
                legend = "none",
                ggtheme = ggpubr::theme_pubr(), 
                x.text.angle = 0, palette = "Paired"
                )


#######



## Using patchwork
archR_simdata_quant_results_plot_woLabels <-
((archR_ari_pl1 + nplb_ari_pl1 + 

  ((archR_ari_pl2 / nplb_ari_pl2 ) + plot_layout(heights = c(2,1))) + 
    plot_layout(widths = c(3,1,1))) / 
  
  ((archR_time_pl + nplb_time_pl) + plot_layout(widths = c(4,1))) + 
  
  plot_layout(heights = c(3,1)) 
    ) 
  
  
archR_simdata_quant_results_plot  <- archR_simdata_quant_results_plot_woLabels +
  plot_annotation(tag_levels = 'A')



for(devc in use_device){
  use_fname <- file.path("figures", devc, 
                         paste0("archR_manuscript_figure_simdata_quant_results.", devc))
  ggplot2::ggsave(plot = archR_simdata_quant_results_plot, 
                  filename = use_fname, 
                  dpi=300, width=10, height=11, units = "in")
  
}

## Combining quant and qual result plots into one, hence this is commented
# if(knitr::is_html_output()){
#     archR_simdata_quant_results_plot
# }
# if(file.exists(use_fname) && !knitr::is_html_output()){
#     knitr::include_graphics(use_fname)
# }
    
```



```{r simdata-results, fig.align="center", out.height="19cm", message=FALSE, warning=FALSE, dpi=500, fig.cap="Assessment of performance of seqArchR on simulated data. (\\textbf{A}) Adjusted Rand index (ARI) attained by seqArchR with various parameter combinations on simulated data (1000 sequences) (data described in Table \\ref{tab:table-simulated-data}). Bar heights represent average over ten runs. Experiments are performed for various chunk sizes, mutation rates (m) and number of mutated motif positions (p). (\\textbf{B}) ARI attained by NPLB for 1000 sequences setting. (\\textbf{C}, \\textbf{D}) ARI attained by seqArchR and NPLB respectively for two scaled up versions of the simulated data: 5000 and 10000 sequences. See main text for more details. (\\textbf{E}, \\textbf{F}) Time taken by seqArchR and NPLB respectively to process simulated data (see main text for details)."}

#### m=0.1, p=1 example
 
sim_res <- readRDS(file.path(archR_simdata_results_path, "seed_1448320993", 
    "archR_result_simulateddata_mu0.1_p1_n1000_modSelType_stability_chunkSize_500_bound_1e-08", 
    "archRresult.rds"))

res_clusts <- sim_res$clustSol$clusters

sim_arch_list <- lapply(seq_along(res_clusts), function(x){
  pl <- archR::plot_ggseqlogo_of_seqs(as.character(sim_res$rawSeqs[res_clusts[[x]]]), 
    pos_lab = 1:100, bits_yax = "auto", title = NULL)
  if(x == length(res_clusts)){
    pl + theme(plot.margin = unit(c(0,0.2,-0.5,0.1), "cm"))
  }else{
    pl + theme(plot.margin = unit(c(0,0.2,-1,0.1), "cm"))
  }
})

#### m=0.5, p=3 example
sim_res2 <- readRDS(file.path(archR_simdata_results_path, "seed_1448320993", 
    "archR_result_simulateddata_mu0.5_p3_n1000_modSelType_stability_chunkSize_500_bound_1e-08", 
    "archRresult.rds"))

res_clusts2 <- sim_res2$clustSol$clusters

sim_arch_list2 <- lapply(seq_along(res_clusts2), function(x){
  pl <- archR::plot_ggseqlogo_of_seqs(as.character(sim_res2$rawSeqs[res_clusts2[[x]]]), 
    pos_lab = 1:100, bits_yax = "auto", title = NULL)
  if(x == length(res_clusts2)){
    pl + theme(plot.margin = unit(c(0,0.2,-0.5,0.1), "cm"))
  }else{
    pl + theme(plot.margin = unit(c(0,0.2,-1,0.1), "cm"))
  }
})

## Using cowplot

get_hline_ycoord <- function(res_clusts){
    length(unlist(res_clusts)) - cumsum( unlist(lapply(res_clusts, length)) ) 
}

## for top row
## change once, reflects for each plot of the top row
use_mar <- c(4, 3, 1, 0.2)
use_mgp <- c(1.4, 0.3, 0)
use_cex.lab <- 0.8
use_cex.axis <- 0.8
use_tcl <- NA

r1c1 <- function(){
  par(
    mar = use_mar, mgp = use_mgp,
    cex.lab = use_cex.lab, cex.axis = use_cex.axis, tcl = use_tcl
  )
  archR::viz_seqs_acgt_mat_from_seqs(as.character(sim_res$rawSeqs), 
      pos_lab = 1:100,
      xt_freq = 25, yt_freq = 200)
}

r1c2 <- function(){
  par(
    mar = use_mar, mgp = use_mgp,
    cex.lab = use_cex.lab, cex.axis = use_cex.axis, tcl = use_tcl
  )
  archR::viz_seqs_acgt_mat_from_seqs(as.character(sim_res$rawSeqs[unlist(res_clusts)]), 
      pos_lab = 1:100,
      xt_freq = 25, yt_freq = 100)
  abline(h = get_hline_ycoord(res_clusts), lwd = 2, col = "white")
}

r1c3 <- function(){
  par(
    mar = use_mar, mgp = use_mgp,
    cex.lab = use_cex.lab, cex.axis = use_cex.axis, tcl = use_tcl
  )
  archR::viz_seqs_acgt_mat_from_seqs(as.character(sim_res2$rawSeqs[unlist(res_clusts2)]), 
      pos_lab = 1:100,
      xt_freq = 25, yt_freq = 100)
  abline(h = get_hline_ycoord(res_clusts2), lwd = 2, col = "white")
}

top_row <- cowplot::plot_grid(r1c1, r1c2, r1c3, byrow = TRUE, nrow = 1,
                              labels = c('G', 'H', 'I'), hjust = c(-1, -1, -3),
                              label_fontface = 'plain', label_size = 15
                              ) 

r2c1 <- cowplot::plot_grid(plotlist = sim_arch_list, ncol = 1, rel_heights = c(1,1,1,1.4))
r2c2 <- cowplot::plot_grid(plotlist = sim_arch_list2, ncol = 1, rel_heights = c(1,1,1,1.4))  

bottom_row <- cowplot::plot_grid(r2c1, r2c2, ncol = 2,
                                 labels = c('J', 'K'), vjust = -0.5,
                                 label_fontface = 'plain', label_size = 15
                                 )

archR_simdata_quant_qual_results_plot <- 
  cowplot::plot_grid(archR_simdata_quant_results_plot, 
                     top_row, bottom_row, nrow = 3, byrow = TRUE,
                  rel_heights = c(1, 0.4, 0.2), 
                  rel_widths = c(1, 0.8, 1)
                  )

for(devc in use_device){
  use_fname <- file.path("figures", devc,                          
      paste0("archR_manuscript_figure_simdata_quant_qual_combined_results.", devc))
  
  ggplot2::ggsave(plot = archR_simdata_quant_qual_results_plot,
                  filename = use_fname, 
                  dpi=300, width=12, height=16, units = "in")
  
}


if(knitr::is_html_output()){
    # archR_simdata_quant_qual_results_plot
    archR_simdata_quant_results_plot
}
use_fname <- file.path("figures", devc, 
                paste0("archR_manuscript_figure_simdata_quant_results.", devc))
if(file.exists(use_fname) && !knitr::is_html_output()){
    knitr::include_graphics(use_fname)
}

```




## seqArchR identifies fixed elements of promoter sequence architectures in different organisms

In general, when dealing with DNA sequences, different organisms are known to 
present different challenges for sequence analysis methods.
For example, the high \texttt{GC} and especially CpG content in human genomic 
sequence is known to affect approaches using motif over-representation 
[@down2005nestedmica; @roider2009].
More specifically, with respect to gene promoters, there is diversity of sequence 
elements within promoters of one organism and also between those of different 
organisms. 
Different promoter types have different sequence composition, some of which are 
highly constrained, and some are not.


Drosophila promoter sequences have multitude of well defined sequence motifs 
viz. the \texttt{TATA}-box and various Ohler motifs [@ohler2002computational; 
@ohler2006].
In comparison, promoters of vertebrates, like zebrafish, and 
mammals, like mice and humans, have relatively fewer well defined motifs;
they are also characterised by the stretches of varying sequence compositions, 
e.g., \texttt{CG/GC} dinucleotide content _vs_ \texttt{A/T}-rich `W`-box motifs, 
owing to different architectures (sometimes even intertwined) determining TSS 
selection at play during different stages of development [@haberle2016promoter;
@nepal2020dual].
This diversity of promoter sequence elements lends to diversity in mechanisms 
of transcription regulation.
Thus, computational methods that work for one organism may not work 
off-the-shelf for other organisms.
In the following, we evaluate seqArchR's ability to identify fixed-element 
promoter sequence architectures in fruit fly, zebrafish and humans.



```{r dm-figure-1, out.extra = 'angle=90', out.width='125%', out.height= '125%', fig.align='center', fig.cap="Clusters and architectures identified by seqArchR for \\textbf{\\emph{D. melanogaster}, 2-4h AEL}. Sequence clusters arranged by the median interquantile widths (IQW) of CAGE TCs in seqArchR clusters (shortest on top, broadest at the bottom). From left to right: Box and whisker plots of per-cluster IQWs, TPMs, and PhastCons scores followed by per-cluster sequence logos, and stacked barplots showing proportion of TCs unique/shared between transitions. Sequence logos for histone gene clusters are shown with a grey background. `All' denoting common between all stages. TPM, Tags per million."}
knitr::include_graphics(file.path(dm_result_dir_path[[1]], 
                                  paste0("drosophila_figure_paper_", dm_sample_names[1], ".pdf")))
```


```{r dm-figure-2, out.extra = 'angle=90', out.width='125%', out.height= '125%', fig.align='center', fig.cap="Clusters and architectures identified by seqArchR for \\textbf{\\emph{D. melanogaster}, 6-8h AEL}. Sequence clusters arranged by the median interquantile widths (IQW) of CAGE TCs in seqArchR clusters (shortest on top, broadest at the bottom). From left to right: Box and whisker plots of per-cluster IQWs, TPMs, and PhastCons scores followed by per-cluster sequence logos, and stacked barplots showing proportion of TCs unique/shared between transitions. Sequence logos for histone gene clusters are shown with a grey background. `All' denoting common between all stages. TPM, Tags per million."}
knitr::include_graphics(file.path(dm_result_dir_path[[2]], 
                                  paste0("drosophila_figure_paper_", dm_sample_names[2], ".pdf")))
```


```{r dm-figure-3, out.extra = 'angle=90', out.width='125%', out.height= '125%', fig.align='center', fig.cap="Clusters and architectures identified by seqArchR for \\textbf{\\emph{D. melanogaster}, 10-12h AEL}. Sequence clusters arranged by the median interquantile widths (IQW) of CAGE TCs in seqArchR clusters (shortest on top, broadest at the bottom). From left to right: Box and whisker plots of per-cluster IQWs, TPMs, and PhastCons scores followed by per-cluster sequence logos, and stacked barplots showing proportion of TCs unique/shared between transitions. Sequence logos for histone gene clusters are shown with a grey background. `All' denoting common between all stages. TPM, Tags per million."}
knitr::include_graphics(file.path(dm_result_dir_path[[3]], 
                                  paste0("drosophila_figure_paper_", dm_sample_names[3], ".pdf")))
``` 




### Fruit fly (*Drosophila melanogaster*) {#dm-results}


@schor2017promoter produced high resolution CAGE data and studied the effect of 
natural genetic variation on transcription strength and distribution of 
transcription start sites in gene promoters at different developmental stages 
in Drosophila.
From the various samples available, we randomly selected one sample `RAL28' 
(this number identifies the D. melanogaster line from Drosophila melanogaster 
Genetic Reference Panel (DGRP) that were used by Schor et al.) to help 
demonstrate seqArchR's efficacy.

Specifically, we used seqArchR to identify promoter architectures at three 
stages of transition during embryogenesis \textemdash\space 2-4 hours (h), 
6-8h, and 10-12h after egg laying (AEL).
Each stage was analysed with seqArchR separately.
We used 91 bp-long sequences \textemdash\space 45 bp upstream and downstream of 
the TSS \textemdash\space as core promoter sequences.
Figures \@ref(fig:dm-figure-1), \@ref(fig:dm-figure-2), and 
\@ref(fig:dm-figure-3) show the sequence logos of architectures of all 
clusters identified for the three transitions respectively. 
For brevity, we use the following shorthand notation to refer to clusters at 
any transition: CnX or CnY or CnZ. 
Here, 'C' stands for cluster, n, the cluster number, and letters X, Y and Z 
denote the three transitions 2-4h, 6-8h, and 10-12h AEL.
A single cluster, say cluster 3 at 2-4h, is named C3X, and multiple clusters, 
say clusters 2-6 or clusters 3 and 5 at 10-12h are named C2-6Z or C3,5Z 
respectively.
Unless specified otherwise, all clusters are arranged in ascending order of 
their median interquantile width (IQW).


#### Motifs

seqArchR identifies several of the known core promoter elements in Drosophila. 
These are directional motifs with a precise or variable positioning within 
promoters \textendash\space abbreviated as DMp and DMv respectively.
The identified DMp motifs include the `TATA`-box, 
initiator (Inr, canonical and non-canonical), the downstream promoter element (DPE). 
Among DMv motifs, seqArchR identifies Ohler motifs 1 and 6, and another motif 
named DMv5 (@ohler2002computational; @fitzgerald2006comparative).
seqArchR also identifies some of the non-directional motifs (NDMs) reported 
by @fitzgerald2006comparative such as the downstream regulatory element (DRE) or 
Ohler motif 2 (NDM4).
The non-directional sequence elements are visible in the sequence logos of the 
raw clusters from seqArchR which were manually combined to obtain the final 
clusters as seen in the main figures (see Supplementary Figures 6, 7 and 8).


`TATA` and DPE sequence elements are known to mutually exclusively pair with 
the Inr for architectures characteristic of tissue-specific and developmental 
genes respectively.
Very early in the embryogenesis, at 2-4h, we do not see any clusters with a 
`TATA`-box except for the histone gene clusters.
C6X with Inr+DPE architecture and C1Y with Inr+DPE1 (variant of DPE) architecture 
have a very short median IQW denoting highly focused transcription.
C8Y combines the cognate initiator, Inr1, with the DPE.
At 10-12h, in addition to the DPE architectures (C2Z and C4Z), we see the 
`TATA`-box architecture in C3Z where the \texttt{TATA}-box combines with a 
canonical dinucleotide initiator `C`$_{-1}$`A`$_{+1}$.
In C8Z, there are two initiators, the canonical `CA` and full Inr, their 
`A`$_{+1}$ 3nt apart.
We speculate that the canonical `CA` is paired with the `TATA`-box at $-31$ bp 
from the TSS, and the Inr at $+3$ bp position is paired with a weaker DPE 
which is also correspondingly shifted by 3 bp.
C11Z has an Inr with a `TATA`-like sequence at $-30$. 
Many CTSSs in these tissue-specific architectures (with TATA and DPE) are unique 
to 10-12h (relatively higher percentages of sites unique to a transition point) 
denoting that these start sites are not active earlier.

`TGT` motif at initiator position is found in all timepoints \textemdash\space 
C18,20X and C18Y and C20Z.
`TTAGT` is identified as the initiator in C9Z. 
A variant of this initiator with a weaker `G` is found  in C10X, C12Y.
C9Z has very different gene ontology (GO) terms enriched in comparison to C10X 
and C12Y, both of which share atleast a few terms as seen in Figure 
\@ref(fig:go-terms-comparison-dm). See Supplementary Figure 10 for the top-10 
enriched GO terms.

```{r go-terms-comparison-dm, fig.align='center', fig.cap="Visualisation of GO terms enriched for various clusters at different developmental stages of Drosophila melanogaster. Top-5 enriched GO terms are shown for each comparison.", out.height="75%"}

knitr::include_graphics(file.path(dm_result_dir_path[[3]], 
                                  paste0("combined_top5_GO_term_enrichments_Drosophila_clusters.pdf")))


```


#### Functional specialisation of clusters

As expected, `TATA`-box and DPE are among the top-ranked focused architectures 
(clusters with lower median IQW) overall.
Genes responsible for translational machinery, such as the ribosomal protein 
genes, exhibit the polypyrimidine initiator, `TCT`, architecture.
These genes are known to have ubiquitously high expression. 
The corresponding clusters across stages, namely C7X, C11Y and C10Z, 
have the highest median TPM values while still being relatively sharp.



##### Detection of recently expanded gene families
seqArchR is capable of identifying promoters associated with histone genes 
which are known to have sequence repeats. 
These are identified across all timepoints.
Specifically, C1-5X, C2-4Y and C6-7Y, and C1Z and C5-7Z are enriched for 
histone genes His1, His2A/2B and His4/4r and His3.
This explains the very high information content of the sequence logo throughout 
the core promoter region.
All CTSSs of His1 gene, namely C2X, C3Y and C6Z, have TSSs in 5' UTRs.
Clusters C4X, C6Y, and C5Z, all clusters of His2B gene CTSSs, are contaminated
with some non-histone ones
(See section 3.2 in the Supplementary text for more details on this contamination).



##### Novel Observations
seqArchR also detects a novel positional motif consisting of a polythymine stretch, 
reminiscent of  `T`-blocks (with >= 3`T`s) common in C. elegans core promoters 
(@tblocks2011yanai). 
We observe such T-blocks at transitions 2-4h and 10-12h AEL. 
Specifically, C11X and C13Z, both contain `T`-blocks flanking the CTSS of core 
promoters in this cluster. 
`T`-stretches have been reported in human genome, too, where they were shown to 
be able to bind TBP (@tatapolyT).


C15X, C14Y and C15Z, all have `A`/`T` enrichment at 30 to 45 bp downstream of 
the CTSS, with a marked increase precisely at 35 bp.
The GO terms enriched for these clusters are also similar \textemdash\space 
mainly including splicing- and catabolic/metabolic process-related terms 
(Figure \@ref(fig:go-terms-comparison-dm)E and Supplementary Figure 10).
We speculate that this architecture is related to the reported pausing of the 
RNA PolII between $+20$ bp and $+50$ bp with the centre at $+35$ bp from the TSS 
(@mavrich2008nucleosome).
This arrangement places the front end of PolII just 10 bp away from the $+1$ 
nucleosome border and enables their contact.
@kwak2013precise also report on promoter-proximal PolII pausing (termed 
focused-proximal pausing) at 35-40 bp downstream of the TSS.
Indeed, the gene RpS7, an example proximal-pausing gene from @kwak2013precise 
is present in these clusters.



#### Successions of motif use in Drosophila development
When comparing promoter sequence architectures prevalent at different stages of 
Drosophila development, some trends immediately emerge.
While the DPE is observed from 2-4h and beyond (C6X, C2Y and C4Z), a canonical 
`TATA`-box only appears at 10-12h stage (C3Z).
This confirms the idea that the DPE is associated with developmental regulation, 
and `TATA` with tissue-specific expression in differentiated cell types 
[@engstrom2007genomic].
Unlike `TATA`, DPE, `TTT` and initiator, other motifs identified by 
over-representation (e.g., Ohler motif 1/6, DRE, etc.) from 
@ohler2002computational and @fitzgerald2006comparative do not occur at fixed 
distance from the TSS.


For completeness, we also analysed Drosophila melanogaster promoter sequences 
from modENCODE [@chen2014comparative] with seqArchR to compare its performance 
on real promoter sequences with that of the current state-of-the-art approach, 
NPLB [@mitra2016nplb].
seqArchR identifies all architectures including variations of the `TATA`-box 
architectures with their precise positional changes.
While seqArchR was run for 5 iterations for this data, it identifies a 
comparable set of clusters/architectures after three 
iterations and just 38 minutes compared to the 600 minutes taken by NPLB (for 
its first pass).
See additional details in Supplementary Information, section 2.




### Zebrafish (*Danio rerio*) {#zf-results}

@nepal2013dynamic produced the first single nucleotide resolution transcriptome
(CAGE) data for a vertebrate, zebrafish, across different stages of 
embryonic development.
We used seqArchR to identify different promoter architectures in zebrafish at 
three different developmental stages, from early to late, namely 64 cells 
(early maternal), 30% Epiboly/Dome (late maternal/early transition), and the 
Prim-6 stage (late zygotic).
Promoter sequences at each stage were independently analysed with seqArchR to 
identify clusters and architectures.
We use the same naming scheme, as used for D. melanogaster, for referring to 
clusters identified in zebrafish, except X, Y and Z here denote the three 
developmental stages of zebrafish analysed in this study.

Figures \@ref(fig:zf-figure-1) and 
\@ref(fig:zf-figure-2) show the clusters identified for 64 cells stage, 
30% Epiboly and Prim-6 stage respectively.
At 64 cells stage, which is only 2 hours post fertilisation (hpf), all 
architectures contain the canonical pyrimidine-purine initiator with an 
upstream `W`-box \textemdash\space a `TATA`-like element, but more degenerate 
than canonical `TATA` box.
This is the architecture used for maternal transcription in the oocyte 
[@haberle2014shifting]. 


```{r zf-figure-1, out.extra = 'angle=90', out.height= '105%', out.width='100%', fig.align='center', fig.cap="Clusters and architectures identified by seqArchR for \\textbf{(A) 64 cells} stage and \\textbf{(B) 30\\% Epiboly/Dome stage}. Sequence clusters arranged by the median interquantile widths (IQW) of CAGE TCs in seqArchR clusters (shortest on top, broadest at the bottom). Each panel, from left to right: Box and whisker plots of per-cluster IQWs, and TPMs followed by per-cluster sequence logos, and stacked barplots showing proportion of different genomic annotations. TPM, Tags per million."}

knitr::include_graphics(file.path(zf_result_dir_path[[3]],
                                  paste0("zebrafish_figure_paper_combined_stage_1and2.pdf")))


```


```{r zf-figure-2, out.extra = 'angle=90', out.width='105%', out.height= '105%', fig.align='center', eval=TRUE, echo=FALSE, fig.cap="Clusters and architectures identified by seqArchR for \\textbf{Prim-6 stage} of zebrafish development. Sequence clusters arranged by the median interquantile widths (IQW) of CAGE TCs in seqArchR clusters (shortest on top, broadest at the bottom). From left to right: Box and whisker plots of per-cluster IQWs, and TPMs followed by per-cluster sequence logos, and stacked barplots showing proportion of different genomic annotations. For cluster C10Z, an additional zoomed-in view of the section from 45 bp to 150 bp downstream is shown. TPM, Tags per million."}

knitr::include_graphics(file.path(zf_result_dir_path[[3]],
                                  paste0("zebrafish_figure_paper_", zf_sample_names[3], ".pdf")))

```




#### Motifs
In comparison to Drosophila, core promoters of zebrafish (and other vertebrates) 
do not possess as much variety of sequence motifs at fixed distance from the 
dominant TSS.
In zebrafish, across developmental stages, genome-wide transcription initiation 
occurs at `Y`$_{-1}$`R`$_{+1}$ (see Figure \@ref(fig:zf-figure-1), 
and \@ref(fig:zf-figure-2); @nepal2013dynamic).
Motif `GGG`$_{+1}$ is seen in some clusters (C3,5Z) where majority of the CTSS 
are located in non-promoter regions, especially exons.
Zebrafish core promoter sequences are known to harbour the `W`-box motif in 
lieu of the `TATA`-box w.r.t. position (~30 bp upstream of TSS) and function 
(directing transcription initiation from a fixed position). 
This is characteristic of maternal architectures.
While some clusters at 64 cells and 30\% Epiboly/Dome stage, namely C1,6X and 
C5-7Y, have enrichment of `WW` downstream starting at ca. 50 bp, this signal is 
not periodic.
In comparison, cluster C11Z, the broadest among all clusters at the Prim-6 stage, 
exhibits a 10 bp periodicity in dinucleotide `WW`.
This periodicity is known to be an intrinsic property of promoters with a 
well positioned `+1` nucleosome downstream following the nucleosome free 
region (NFR) spanning the  promoter [@dreos2016influence].
This is characteristic of broad promoter architectures where the downstream 
nucleosome creates a so called `catchment' area for transcription initiation 
[@haberle2016promoter]. 
As recently reported (Baranasic et al., 2022), the chromatin organisation of 
promoters between 30%-Epiboly/Dome and Prim-6 stages remains robust.


#### Functional specialisation of clusters
Even though more than 90% of the stable RNA in 30% Epiboly/Dome stage is still 
maternally inherited [@haberle2014shifting], seqArchR already detects 
zygotically activated promoters and their specific architectures. 
Notably, it detects the promoters of the first wave of zygotic genome 
activation (ZGA), primarily coming from a dedicated gene cluster on chromosome 
4 (clusters C3,9-10Y) (Figure \@ref(fig:zf-add-figure)A). 
Cluster C9Y belongs to genes encoding mir-430, which are present in hundreds of 
moderately divergent copies [@yavor2023mir], and whose miRNA product 
specifically targets maternal RNAs for degradation [@ZFmir-430-1; @ZFmir-430-2]. 
The ability to detect these architectures in a sample dominated by maternal 
RNA shows that seqArchR is a powerful tool for characterising regulatory 
transitions at the promoter level. 

Across all clusters at 64 cells stage, the proportion of promoters that are 
exclusively used at this stage (stage-specific) is very low (Supplementary 
Figure 14).
This behaviour continues into the 30\% Epiboly/Dome stage with exceptions of 
clusters with promoters from chromosome 4 (C3,9-10Y).
Contrastingly, at the Prim-6 stage, many new 
stage-specific promoters get introduced (average 25\%).



##### Gene ontology of main clusters


The GO terms enriched for selected clusters are shown in Figure 
\@ref(fig:zf-add-figure)B. 
In the figure, the clusters across different developmental stages are grouped 
based on the attributes of their architectures. 
The set of clusters on the left (group with red bar along the horizontal axis) 
is clusters with no downstream enrichment of `W` (`A/T`) signal _vs_ that on the 
right (blue bar) is those with downstream `W` signal.
The left group has distinctive enrichment for development-related GO terms.
Three clusters at stage Prim-6, C7-9Z, which are part of the left group here, 
additionally lack the upstream `W`-box unlike other clusters in this group. 
These clusters are enriched for different GO terms (Figure \@ref(fig:zf-add-figure)B).
The GO terms exclusively enriched for cluster C8Y are related to arresting 
of process of fertilisation \textemdash\space  positive regulation of 
cellular component organisation and negative regulation of fertilisation 
(Supplementary Figure 15).


```{r zf-add-figure, fig.cap="\\textbf{(A)} Genomic locations of per-cluster promoters in Zebrafish developmental stages 64 cells, 30\\% Epiboly/Dome and Prim-6. \\textbf{(B)} Comparison of top-5 enriched GO terms for different clusters per stage of zebrafish development. Clusters grouped by architecture attributes: \\colorbox{red!40}{absence} \\emph{vs} \\colorbox{blue!40}{presence} of downstream enrichment of \\texttt{W (A/T)} signal.", out.extra = 'angle=0', out.width="100%", fig.align='center'}

knitr::include_graphics(file.path(zf_result_dir_path[[3]],  "combined_top5_GO_mix_clusts_chr4_prop.pdf"))

```


#### Motif successions in early zebrafish development
Comparing promoter architectures in early to later stages, the shift from 
maternal `W`-box\textendash based architectures is evident.
In the 30\% Epiboly/Dome stage, there are clear indications of emerging promoter 
sequences that do not conform to the maternal architecture seen in the 64 cells 
stage.
By Prim-6 stage, about half of the promoter sequences harbour the 
architecture where dispersed transcription initiation is orchestrated by the 
first downstream nucleosome (C10Z).



### Human (*Homo sapiens*) {#human-results}


```{r hg-figure-1, out.extra = 'angle=90', out.width='125%', out.height= '125%', fig.align='center', fig.cap="Clusters and architectures identified by seqArchR for CAGE-derived \\textbf{core promoter sequences from cell lines and tissues} of H. sapiens. Sequence clusters arranged by the median interquantile widths (IQW) of CAGE TCs in seqArchR clusters (shortest on top, broadest at the bottom). From left to right: Box and whisker plots of per-cluster IQWs, TPMs, and tissue specificity scores ($\\tau$) followed by per-cluster sequence logos, and stacked barplots showing proportion of different genomic annotations. TC, tag clusters. TPM, tags per million."}

knitr::include_graphics(file.path(hg_result_dir_path[[1]],
                                  paste0("human_figure_paper_", hg_sample_names[1], ".pdf")))

```

```{r extra-fig-1, fig.align = 'center', fig.cap="Additional analyses of seqArchR clusters for \\textbf{H. sapiens}. \\textbf{(A)} Per-cluster proportions of ribosomal and other (non-ribosomal) genes (left) and proportions of dual initiation and non-dual initiation promoters (right). \\textbf{(B)} Proportion of housekeeping (HK) genes in each cluster with clusters arranged in ascending order of \\% HK genes (bottom) and absolute numbers (top). seqArchR cluster names correspond to those from the previous figure."}

knitr::include_graphics(file.path(hg_result_dir_path[[1]],
                                  paste0("extra_figure_human", ".pdf")))

```

Human promoter sequences are known to be challenging for sequence analysis 
methods owing to their sequence heterogeneity and, in the case of majority of 
protein-coding genes, high GC and CpG percentages. 
To test seqArchR on human core promoter sequences, we gathered CAGE data for 
different human cell lines and tissues from ENCODE. 
They were pooled by merging CAGE data for all cell lines (see Supplementary 
Figure 14). 
Using a threshold of 1 TPM, about 9500 promoter sequences were analysed using 
seqArchR. 


#### Representative promoter windows

First, we tried the same sequence window as in zebrafish ($-45,+150$) 
because we expected from the FANTOM5 analysis (supplementary figure 13 in 
@andersson2014atlas) that a majority of CpG-island overlapping and broad promoters will 
have a stable position of $+1$ nucleosome and an underlying nucleosome positioning
signal, albeit weaker than in zebrafish.
Keeping the bound value moderately stringent at $10^{-6}$, the analysis
separated architectures that had a `TATA`-like signal ca. $30$ bp upstream from 
the `CG`-rich architectures.
The inability to identify a more stringer, canonical `TATA` signal as a separate 
class is possibly because of the dominant influence of the downstream flanking 
sequence in the chosen window.
To try to prevent drowning of this TSS-determining motif architecture, we 
repeated the analysis using flanks of $-55$ bp and $+5$ bp around the dominant 
TSS.
Figure \@ref(fig:hg-figure-1) shows the clusters and corresponding architecture 
sequence logos. 
Only `r tata=48+92+88+69;tata` of these, which is 
`r tata_perc = (tata/7000)*100; round(tata_perc, digits=2)`\% of all the 
core promoter sequences, are observed to have the \texttt{TATA} motifs 
(Figure \@ref(fig:hg-figure-1), clusters C1 and C3-C5).
We speculate that the reason for this low percentage could be due to the way 
the set of promoters was obtained: by pooling CAGE signal from many different 
tissue and cell types where signal from tissue-specific genes is diluted, unlike 
that of broadly expressed genes, bringing a higher proportion of the later 
under the 1TPM threshold. 
We used the $\tau$ values as a measure of tissue specificity of genes 
[@yanai2005genome]. 
Of these, clusters C1, C3 and C4 have a higher median tissue specificity 
score than the rest of the clusters. 
The proportion of housekeeping genes is also extremely low in these clusters 
(Figure \@ref(fig:extra-fig-1)B). 
The top-5 enriched GO terms for each cluster (excluding C2 and C7-8, which have
$>25\%$ non-promoter CTSSs) are shown in panel C, Figure \@ref(fig:extra-fig-1).
A clear distinction of enriched terms is observed between tissue-specific/focused 
architectures and broader architectures.
Focal adhesion, a cellular component GO term is enriched across all broader 
architecture clusters (C9-14).


The `TCT` architecture seen in C6 contains the largest percentage of ribosomal 
protein genes (Figure \@ref(fig:extra-fig-1)A). 
They are also seen in the `TATA` architecture clusters, 
albeit much fewer in proportion. 
This is not surprising given that earlier studies have reported that the 
promoters of these genes are known to harbour `TATA` or `TATA`-like sequences 
[@RPGyoshihama2002human; @RPG2006characteristicsHuman; @RPG2011basalmetazoans].
@nepal2020dual reported on dual initiation promoters, where `YC` initiation 
is intertwined with canonical initiation (`YR`) in metazoans, and also being 
pervasive in human and Drosophila. 
It was also shown that dual initiation promoters are often broad in shape owing 
to many TSSs in vicinity of each other.
Our observations here for human promoter architectures are in line with this 
(see Figure \@ref(fig:extra-fig-1)B) \textemdash\space proportion of dual 
initiation promoters is higher in clusters with broad promoters, with 
cluster C2 (non-promoter CTSSs) and C6 (`TCT`/`YC` architecture) as exceptions. 

#### Exclusion of initiator sequence and subsequent analysis

We envisaged that, at least in vertebrates, because the initiator is degenerate 
(`YR`), its strength may lead to splitting clusters. 
Thus, to test this importance of the initiator towards the clustering of the 
promoter sequences, we compared the clusters obtained with and without the 
initiator sequence. Refer to Supplementary Figure 12 for the following.
The results show some interesting observations. 
In most of the clusters, a large chunk of promoters stay together in both 
clusterings. 
Examples of such behaviour are clusters C2-5, C8-9, C11-13, C17,19-20 (left 
panel; with initiator). 
These promoters roughly retain their shape-based rankings between the two 
clusterings.
In some clusters, such as C14-16 (with initiator) where the flanks show slight 
`GC` enrichment, a good chunk of these promoters get distributed across some 
narrow and broad architecture clusters, implying that they clustered together 
due to the initiator sequence `TG`.



# Discussion {#discussion}

Promoter sequence architectures are much more than just individual sequence 
motifs lying anywhere within the promoter region. 
In some instances, these sequence motifs must occur in tandem, at fixed 
(relative) distances from one another.
In addition to motifs, the sequence composition also plays an important role.
This gives rise to a diverse set of sequence architectures even within a single 
species.
Organisms are known to use different core promoter grammars as may be suitable 
in different scenarios.
They can be the different stages of embryogenesis of an organism, e.g., early 
_vs_ late, or for different functions, e.g., tissue specificity _vs_ housekeeping 
function.
Moreover, there is diversity even across organisms.
Therefore, there is a need for *de novo* approaches to explore and study the 
diverse promoter sequence architectures across species and organisms.


We present seqArchR, an approach for *de novo* identification of sequence 
elements based on non negative matrix fatorisation.
seqArchR can be used for studying promoter sequences aligned by the position 
of their TSSs.
seqArchR's chunking-based iterative algorithm can handle scenarios of complex 
interactions 
between sequence (motif) elements.
Experiments on simulated DNA sequences show its ability to identify strong 
and/or weak artificially planted sequence elements.
seqArchR is recognised as much faster (ca. $20-30$x) than No Promoter Left 
Behind, the current state-of-the-art approach for *de novo* identification of 
promoter sequence architectures.


Different organisms are known to pose different challenges for sequence analysis. 
For example, Drosophila core promoter sequences are known to harbour multiple 
motifs with high information content embedded in otherwise low information regions. 
The core promoter motifs \texttt{TATA}-box, DPE, Ohler motif 5 etc. are some 
such examples.
In comparison, other organisms lack such motifs.
For example, Zebrafish have the `W`-box upstream of the TSS, and the 
`WW` periodicity downstream demarcating nucleosome positions [@haberle2014shifting]. 
Only a small percentage of core promoters in H. sapiens have a 
\texttt{TATA}-box, while the majority of their promoter sequences have a higher 
\texttt{GC} content throughout.
seqArchR is able to detect all types of motifs and non-motifs\textendash 
based sequence elements.
We showed this by processing promoter sequences across three different 
organisms, namely fruit fly, zebrafish, and human.
Depending on the nature of sequence elements, some parameters, namely the 
stability bound, number of iterations, and per-iteration collation decision, 
for seqArchR need to be set appropriately.
Adjusting the *stability bound* value enables adjusting the sensitivity of seqArchR to 
identify even weak sequence features that may be enriched in very few sequences.
As discussed in the [Methods](#archR-algorithm) section, a good choice of number of 
iterations and the per-iteration collations can help identify overall 
defragmented clusters still keeping clusters with minor positional variations 
separate.
Indeed, the ability to tune seqArchR in this way makes it work seamlessly 
across organisms.



seqArchR is developed to discover promoter sequence architectures which are 
tied to the precise transcription initiation sites identified with CAGE experiments.
Therefore, it treats absolute position information of sequence features as 
an important characteristic of any architecture.
We note that for data from some molecular biology experiments, this 
requirement to encode and/or discern precise position information of 
features may not be as important.
This is due to two reasons.
First, because the experiment itself may not be high-resolution (single 
nucleotide), and, second, only relative positions are important for the biology 
being studied.
For instance, ChIP-seq is not high-resolution and relative positions of 
sequence features (transcription factor binding sites) are important.
We envisage that ChIP-exo or ChIP-nexus data, which gives high-resolution 
information of TFBSs, can be processed with seqArchR.

\textbf{Is seqArchR another motif finder?}
seqArchR has some similarities and distinctions w.r.t. any motif finding 
method.
Because seqArchR analyses sequences that can contain enriched motifs, it is 
able to identify motifs, albeit positioned ones, like any other motif finder.
But, seqArchR uses the complete sequence to directly infer clusters of 
sequences with common features.
This is affected by the composition of the whole sequence, including the 
flanks of the motif, if present.
This is unlike any motif finding approach, which looks for statistically 
enriched words of pre-determined size (motif length) (usually 6-15 bp and only 
seldom larger).
Another important distinction is that seqArchR does not require/use any 
background (sequence) information as any motif finder does.
This limits seqArchR's ability to find motifs enriched in set A _vs_ set B 
which is a common scenario where motif finders are used (i.e., differential motif 
enrichment).
In this case, seqArchR can still be used to independently analyse set A and set
B and then compare the identified features.
In a nutshell, seqArchR is a clustering approach, where as typical motif 
finders can be considered as classification-based.



seqArchR can be improved in the following ways.
It currently uses the same set bound value for every chunk/cluster in 
every iteration. 
It may be beneficial to extend this to adaptively change the bound value per 
chunk/cluster in any iteration.
In terms of its algorithm, the approach for identifying overfit clusters can 
be improved.
Furthermore, candidate approaches to automate the 
decision to collate after every iteration can be explored.
Also, the current approach for collation can be improved to reduce manual 
intervention.
In its current form, seqArchR only reports complete profiles of the same 
length as the input sequences.
We want to extend it to be able to extract motifs from complete profiles 
similar to TF-MoDISco [@tfmodisco].

 


# Conclusions {#conclusions}

seqArchR is a *de novo* approach for discovering and clustering promoter sequence 
architectures.
With seqArchR, we demonstrated the ability of NMF to simultaneously discover 
clusters of sequences with diverse positioned sequence motifs.
seqArchR seamlessly identifies promoter architectures across organisms which 
are characterised by sequence motifs or the underlying sequence compositions.
It is much faster than the current state-of-the-art, NPLB.
seqArchR is made available as an R/Bioconductor package [http://www.bioconductor.org/packages/seqArchR](http://www.bioconductor.org/packages/seqArchR) 
and a website accessible at [https://snikumbh.github.io/seqArchR](https://snikumbh.github.io/seqArchR).



# Materials

All experimental CAGE datasets analysed in this study were obtained and 
processed as follows.
CAGE data for embryonic developmental stages in Drosophila melanogaster were
obtained from @schor2017promoter.
Out of different strains/samples available, sample 'RAL_28' was randomly chosen.
CAGE data for three transitions in embryonic development, namely 2-4h, 6-8h and
10-12h after egg laying, 
was processed using CAGEr v2.0.2. See [this](link here) Rmarkdown document for 
more details.

CAGE data for the embryonic developmental stages in Zebrafish were obtained 
from @nepal2013dynamic. 
Out of all stages ranging over maternal to zygotic transition, we selected 
three, namely 64 cells, 30% Epiboly or Dome, and Prim-6 for this study.
These were processed using CAGEr v1.20. See [this](link here) Rmarkdown 
document for more details.

For Homo sapiens, all cell lines-specific CAGE data was obtained from ENCODE 
consortium which was made available previously as an R package 
`ENCODEprojectCAGE_1.0.1.tar.gz` available at [http://promshift.genereg.net/CAGEr/PackageSource/](http://promshift.genereg.net/CAGEr/PackageSource/).
For this study, data from all cell lines was merged and processed as described 
in [this](link here) 
Rmarkdown document.
Library sizes of all cell lines that are included are shown in Supplementary 
Figure 4.

The list of housekeeping genes in humans was obtained from @eisenberg2013human.
The $\tau$ index, first introduced by @yanai2005genome, for all human genes 
were obtained from @tau2021ageing.
@tau2021ageing used the GTEx data to compute the $\tau$ index of 
tissue-specificity for all human genes.
A value closer to zero indicates equally expressed across included tissues in 
the GTEx data and a value closer to $1$ indicates expressed in one tissue.

The list of dual initiation promoters in humans was obtained from @nepal2020dual.


# Methods {#methods}

NMF seeks a low-rank decomposition of any given matrix $V_{p\times n}$ which 
only has non-negative entries.
Here, $n$ is the number of samples and $p$ the number of attributes.
We note that attributes can also be referred to as features and $V$ as data 
matrix.


```{=tex}
\begin{equation}
  V_{p \times n} \approx W_{p \times k} H_{k \times n}\qquad k \ll p;\quad V, W, H >= 0
  (\#eq:nmf)
\end{equation}
```
The matrix $H_{k\times n}$ records the coefficients of the $n$ samples in the $k$-dimensional representation, $k \ll p$.
The matrix $W_{p\times k}$ provides the loadings of the original attributes, $p$, in the $k$-dimensional space.
$W_{p\times k}$ is called the basis matrix (and we refer to each column of $W$ as a basis vector), and $H$ the coefficients matrix.
All entries in $W$ and $H$ are non-negative.
Thus, NMF achieves dimensionality reduction, and can be further used for clustering using information in the coefficients matrix.
Here, it is assumed that the lower-dimensional representation obtained using NMF corresponds to different clusters present in the data, and the clusters are characterised by different groups of attributes.
Due to non-negativity, NMF lends an intuitive parts-based representation in many applications where negative values do not have an intuitive meaning.

The basic idea of seqArchR is to process a given set of sequences with NMF.
These sequences are represented in the form of a one-hot encoded matrix, whose each column corresponds to a sequence and the sequence features are along its rows.
In the following, we describe in detail the input to seqArchR and its chunking-based iterative algorithm. Together with it, information on the relavent parameters from the corresponding R package seqArchR is also given.

## Input
The input to seqArchR is a matrix of one-hot encoded representation of DNA sequences of same length, $L$.
For each DNA sequence, its mononucleotide profile is one-hot encoded as follows.
The four channels in its one-hot encoding \textemdash\space due to DNA alphabet, $\Sigma = \{\texttt{A,C,G,T}\}$ \textemdash\space are concatenated to produce a single column vector representation per sequence.
The size of this vector is four times the length of the sequence itself.
Thus, for an input matrix representing mononucleotide profiles of $n$ sequences, its dimensions are $(4L \times n)$.
All entries in this matrix are non-negative, or more specifically, $0$ or $1$.

Similarly, the dinucleotide profile of a sequence can be one-hot encoded.
The input matrix dimensions in this case are $(4^2L \times n)$.
It is known that dinucleotide profiles of DNA sequences hold more information than mononucleotide profiles (`TO-DO/BL`: citation).
All results reported in this article are obtained using dinucleotide profiles.

As shown in Figure \@ref(fig:archR-algo-figure)B, in the decomposition of the one-hot encoded input matrix $V_{p\times n}$ to $W_{p\times k}$ and $H_{k\times n}$, $W_{p\times k}$ assigns scores to each (di)nucleotide at a given position.
Each column of $W_{p\times k}$, a single basis vector, captures combinations of various nucleotides at characteristic positions in sequences and can be said to represent an architecture.
As an example, consider architectures I-IV in Figure \@ref(fig:archR-algo-figure)B.
All important features ((di)nucleotide-position pairs) are captured at once in a single basis vector of $W$.
$H_{k\times n}$ gives the coefficients for each sequence on all $k$ basis vectors.
These coefficients are relative in nature.
When a particular sequence has features characterised by basis vector A but not those by basis vector B, it gets assigned a higher coefficient for basis vector A than for B.
This information can be used for soft clustering of sequences.
In this fashion, coefficients for all $n$ sequences can be interpreted.
Thus, NMF can simultaneously learn the complex interdependencies and interactions between sequence features *de novo*, and also provide soft clustering of input sequences.


Given a set of DNA sequences as a FASTA file or a \texttt{Biostrings::DNAStringSet} object [@pages2017biostrings], the R package seqArchR has provision to generate one-hot encoding of mono- as well as dinucleotide profiles of sequences.
This can then be used as an input.

## Algorithm {#archR-algorithm}

seqArchR's algorithm for *de novo* identification of clusters and architectures among the given sequences is depicted pictorially in Figure \@ref(fig:archR-algo-figure)A.
Here, we describe the algorithm in full.


1.  **Chunk the given sequences.** 
Given the input matrix, the total collection of sequences is divided into smaller chunks (subsets).
These *chunks* are of a pre-defined size that can be set by the user (parameter: `chunk_size`).

    Note on terminology: With every chunking operation on a set of sequences, the resulting chunks of sequences are referred to as *inner* chunks, and the parent set of sequences are referred to as the *outer* chunk.
    At the first iteration, we have one outer chunk (the complete set of input sequences), and many inner chunks.
    At subsequent iterations, we expect more than one outer chunks, and one or more inner chunks per outer chunk.

2.  **Independently process each inner chunk with NMF as follows.**

    (a) **Finding the appropriate number of clusters/basis vectors (*model selection*):** The number of basis vectors is given by $k$ in Eq. \@ref(eq:nmf).
        We test a range of values for $k$ (parameters: `k_min` and `k_max`) from which the optimum value is selected based on either of the two quantitative criteria mentioned below (parameter: `mod_sel_type`).

        -   *Model selection using stability of obtained basis vectors*.
            @amariDistPNAS proposed and used the Amari-type distance that for a given value of $k$, measures the instability of the identified basis vectors over multiple runs.
            They termed the set of basis vectors $R$ learned in a single run as a dictionary.
            With $R_{m}$ dictionaries from $m$ runs, the dissimilarity between any two dictionary pairs $R, R'$ is given as

            ```{=tex}
            \begin{equation} 
            diss(R,R') = \frac{1}{2K}\bigg(2K - \sum_{j=1}^{K}\max\limits_{1\leq k\leq K} C_{kj} - \sum_{k=1}^{K}\max\limits_{1\leq j\leq K} C_{kj}\bigg).
            (\#eq:amaridist) 
            \end{equation}
            ```
            The instability of the set of identified $k$ basis vectors is the average dissimilarity of all dictionary pairs over $m$ multiple runs as computed by Eq. \@ref(eq:amaridist).
            @amariDistPNAS used this for studying spatial gene expression patterns.

            With increasing values of $k$, the instability increases. 
            seqArchR sets a bound on the instability.
            The maximal value of $k$ whose instability is lower than the bound is selected.
            In the R package, the `bound` parameter can be set by the user.
            Based on experiments on simulated data and various biological datasets, the recommended value of `bound` is one of $10^{\{-6,-7,-8\}}$.
            Higher powers are more stringent and lower ones more lenient.
            The user can choose values beyond this range as may be suitable for a dataset.
            The default value of *bound* is set to $10^{-6}$ (lenient). 
            The default/recommended number of bootstrapped iterations when using this approach for model selection is 100 (parameter: `n_runs`).

        -   *Model selection using bi-cross-validation.* 
            $r$-fold bi-cross-validation (bi-CV) approach was proposed by @owen2009bi and recently used by @plos-patterns-nmf.
            This approach tests the re-construction accuracy of the NMF solution for each value of $k$ and chooses the one that maximises it.
            Multiple bootstrapped iterations of NMF are performed for each value of $k$.
            The one with the maximum average re-construction accuracy over all bootstrapped iterations is chosen as the best one.
            
            The default/recommended number of folds and bootstrapped iterations when using this approach is 5 and 500 respectively (parameter: `cv_folds` and `n_runs`).
            Note that the multiple runs are performed per-fold. 
            The effective number of iterations thus performed is `r 5*500` per value of $k$.
        
        The stability-based model selection procedure is the recommended approach
        <!-- for aligned sequences [Note: keep this bit if the paper discusses both aliigned and non-aligned seqs, aligned as in aligned by the TSS position] -->
        since it is faster owing to the smaller number of bootstrapped runs required per value of $k$.
        
        <!-- (TODO: If we show results on chip-seq data, we can refer to it here.)  -->
        

    (b) **Obtaining clusters:** 
        We then obtain the final NMF solution from multiple runs using the chosen number of basis vectors.
        <!-- We use scikit-learn module in Python for solving NMF.  -->
        We use the coordinate descent solver [@cdsolver2009] with Nonnegative Double Singular Value Decomposition (NNDSVD) initialisation [@nndsvd2008] which encourages sparsity.
        <!-- We also employed $\ell 1$-regularisation. (With the regularisation parameter set to 'None', setting L1 regularisation doesn't work. But in the current version we are not using this.) -->
        Matrices $W$ and $H$ from the run achieving the best re-construction accuracy are collected.
        The set of sequences in the *inner chunk* is then partitioned into as many clusters as the number of basis vectors using the coefficients in matrix $H$ (Eq. \@ref(eq:nmf)).
        Each sequence is added to the cluster corresponding to the basis vector for which it has the highest coefficient.
    
    (c) **Accounting for overfitting.**
        Either of the model selection approaches can suffer from overfitting.
        This leads to identification of clusters where too many features of only few sequences from the cluster are deemed as important. 
        The coefficients assigned to these few sequences for this architecture are very high in comparison to other sequences in the cluster.
        Thus, the distribution of the coefficients for all sequences in such a cluster shows a large variation.
        seqArchR detects such cases by default and controls for them by re-merging them with the parent cluster.
        For example, if a (parent) cluster is divided into three clusters of which the second is adjudged as an overfit cluster, it is merged back into the first 
        (considered as the parent) of the three clusters.
        <!-- (`TO-DO`: We may have to explain why re-merged with the first cluster.) -->
        <!-- (`TO-DO`: Here it will be important to show that an overfit cluster can occur in between, and that it does not necessarily occur always at the end, i.e., third of three clusters or fifth of five clusters.) -->
        <!-- (`TO-DO`: Showing an example will really help to clearly put forth both the above points.) -->
        
         <!-- Because one bound value is used for each chunk, it may be stringent in case of one chunk but lenient in case of another. -->
         <!-- A basis vector which explains too many features of too few sequences is identified.  --> 
        <!-- As a result, in addition to the selected few sequences that get assigned relatively very high coefficients, some other sequences get assigned coefficients  -->
        <!-- and consequently artificially/erroneously assigning higher coefficients to sequences which -->
        


3.  **Collate similar clusters (if any) from different inner chunks.** Among the clusters obtained from the various *inner* chunks, we collate those that are similar into one.
    This is done by clustering similar basis vectors using hierarchical clustering.
    The distance and agglomeration method used for hierarchical clustering can be set by the user (parameters: `result_dist` and `result_aggl`).
    One set of choices here are applicable for all iterations.
    
    Based on various computational experiments, we recommend using ward.D linkage with Euclidean or correlation distance.
    <!-- Euclidean distance and ward.D linkage have been found to work best when architectures have motifs as in Drosophila promoter sequences (`TODO`: ref appropriate results subsection). -->
    <!-- When the architectures are characterized by stretches of sequence profiles, e.g., `GC`-richness, correlation and complete linkage work better.  -->
    <!-- The second combination is used by default in the package. -->
    Note that using any combination of distance and agglomeration method only affects collation of clusters and does not affect per-chunk cluster/architecture identification by NMF.
     

4.  **Re-iterations:** Designate all obtained clusters as separate *outer chunks*.
    Each *outer chunk* is treated as an independent collection of sequences.
    Steps 1-3 are repeated per outer chunk separately with the following caveat.
    As a result of the collation, if any new outer chunk is much larger than 
    the user-specified `chunk_size`, it is further divided into 'inner' chunks 
    of user-specified `chunk_size` as per step 1 and processed with steps 2 and 3.
    
    The designation of obtained clusters as outer chunks and processing them over 
    the next iteration is handled  automatically. 
    seqArchR performs a user-specified number of iterations in this fashion (parameter: `total_itr`).
    
The clusters identified in the last iteration of seqArchR are collated and 
reported as the final result.
This collation of clusters for final reporting is treated as separate from 
the per-iteration collation.

### Important note on collation of clusters and number of re-iterations
Keeping in mind the exploratory nature of any clustering analysis, the 
decision to collate similar clusters from different chunks (step 3) and 
number of iterations (step 4) have no defaults because they are dependent on 
the nature of the data.

**Suppressing or deferring collation.** 
In addition to the choice of the distance measure and the agglomeration method 
to be used for collating clusters using hierarchical clustering, the 
per-iteration decision to collate clusters is also left to the user 
(parameter: `'set_ocollation'`).
One can suppress collation for a particular iteration or all of the 
specified number of iterations.
For instance, if three iterations are to be performed, the user can choose to 
collate clusters only for the second iteration.
We observed that deferring collation is a useful strategy (data not shown), 
especially in scenarios where the sequence features/motifs have minor 
positional variations.
For example, in Drosophila promoter sequences, where `TATA`-box is 
observed anywhere between 32 to 28 bp upstream of the TSS, deferring collation 
to the second iteration is helpful to detect these minor positional differences 
right away.
In fact, in all experiments reported here, except those with simulated data, 
we defer collation to the second iteration.


**Number of iterations.**
The appropriate number of iterations depends on the nature and number of 
sequence architectures.
It is almost always a good idea to perform at least three iterations or at 
least one iteration 
after a deferred collation exercise when analyzing real promoter sequences.


**Curation of collated clusters.**
Another point to note is that depending on the choices for distance measure and 
agglomeration method, programmatic collation of basis vectors with hierarchical 
clustering can sometimes combine clusters which are perhaps better off separate. 
Indeed many studies require some manual curation when combining position weight 
matrices (PWMs) with hierarchical clustering [@rsatpaper] 
(`TODO`: any other paper to cite?).
Therefore, we recommend that the clusters from the last iteration be left 
uncollated (by appropriately setting the parameter `'set_ocollation'` to `FALSE` 
for the last iteration).
The final output from seqArchR will still be a collation of clusters from the 
last iteration which, as noted above, is treated as independent of per-iteration
collation.
Keeping clusters from the last iteration uncollated provides the user with a 
choice to collate the clusters in a custom manner and curate them when necessary. 
The users can either use built-in functions from seqArchR package or any other 
custom script.
All results reported here follow this strategy (except simulated data results).
Supplementary Figures 6-8, 12-14 and 17 show how this process was carried out 
for raw seqArchR results for D.melanogaster, D.rerio, and H.sapiens respectively.



Thus, seqArchR requires no prior information or specification about the nature 
of architectures.
It can distinguish sequences with motif-based architectures from non-motif-based 
architectures.
If there are motifs, it does not require any information on the expected number 
of motifs or motif lengths or if any of the motifs have gaps (and the size of 
the gaps), or if some of them work cooperatively resulting in a combinatorial 
interplay.


### Rationale for the chunking-based algorithm {#chunking-rationale}


```{r seqArchR-mem-figure, fig.width=5, fig.height=3, fig.align="center", out.height="200px", warning=FALSE, dpi=500, fig.cap="Memory footprint of seqArchR. The maximum resident set sise for different settings of seqArchR is reported."}
## archR memory usage plots

relevant_idx <- dplyr::filter(archR_simdata_results_mem, MaxRSS != -1)$index
mem_melted_df2 <- reshape2::melt(archR_simdata_results_mem[relevant_idx,],
            id.vars = c('chunkSize', 'nSeqs'),
            measure.vars = "MaxRSS"
            )

mem_melted_df2$nSeqsInK <- mem_melted_df2$nSeqs/1000
mem_melted_df2$nSeqsInK <- as.factor(mem_melted_df2$nSeqsInK)
mem_melted_df2$nSeqs <- as.factor(mem_melted_df2$nSeqs)
mem_melted_df2$chunkSizeInK <- mem_melted_df2$chunkSize/1000


archR_mem_pl2 <- ggpubr::ggline(data = mem_melted_df2, 
                x = "chunkSizeInK", y = "value",
                palette = RColorBrewer::brewer.pal(6, "Dark2")[6:4],
                color = "nSeqs", 
                add = c("mean_sd"),
                # error.plot = "pointrange",
                xlab = paste0("Chunk size ", expression('(x 1K)')),
                ylab = "Memory used (MB)",
                legend.title = "# Sequences",
                ylim = c(500, 1300),
                ggtheme = theme_bw(), 
                font.legend = 14,
                font.x = 14,
                font.y = 14,
                font.tickslab = 14, 
                size = 1,
                legend = "top"
                )

for(devc in use_device){
  use_fname <- file.path("figures", devc, 
      paste0("archR_manuscript_figure_simdata_memory_footprint.", devc))
  
  ggplot2::ggsave(plot = archR_mem_pl2, device = devc,
                  filename = use_fname, 
                  dpi=500, width=6, height=5, units = "in")
  
}


# if(knitr::is_html_output()){
#     archR_mem_pl2
# }
# 
# 
# if(file.exists(use_fname) && !knitr::is_html_output()){
#     knitr::include_graphics(use_fname)
# }

#######

```


As discussed in the [Results](#results) section, promoter sequence architectures 
often harbour only slightly varying sequence elements at the same position in 
all sequences.
For example, the canonical YR initiator \textemdash\space case of same length 
motifs overlapping in occurrence positions, and Inr `CA` overlapping with the 
first `CA` of Ohler1 motif in Drosophila. 
A chunking-based approach makes it relatively easier to select the best model 
by alleviating the above mentioned co-occurrence cases and handling fewer 
sequences at once.
Supplementary Figure 4A shows the architectures of clusters, identified among 
Drosophila promoter sequences from modENCODE, using NMF with stability-based 
bound criterion for model selection in the first iteration.
It can be seen that, the first set of identified clusters (across chunks) are 
majorly governed by initiator elements.


Additionally, for larger chunk-sizes, the memory usage increases linearly (see 
Supplementary Figure 2).
Therefore, having the ability to process very large datasets one chunk at a 
time enables processing them with a reasonably smaller memory footprint. 
Qualitatively, a very small chunk size can lead to a fragmented set of final 
clusters.


## Method Availability
seqArchR is made available as an R package with a GNU General Public License 
version 3 (GPL-3).
The website documenting the package is accessible at <https://snikumbh.github.io/seqArchR> 
and the source code at <https://github.com/snikumbh/seqArchR>.
seqArchR was previously named archR (version 0.1.8).
This is also the version used for experiments and results reported in this 
manuscript.
It is deposited at Zenodo with DOI 
[10.5281/zenodo.5055408](https://doi.org/10.5281/zenodo.5055408) 
[@sarvesh_nikumbh_2021_5055408].
Following version 0.1.8, there have been no changes to the algorithm.


# Acknowledgements {.unlisted .unnumbered}

We acknowledge Leonie Roos for processing earlier versions of Drosophila and 
Zebrafish CAGE datasets.
We thank members of the Computational Regulatory Genomics group for discussions 
through out the project and feedback on the manuscript.

# References


