---
author: "Sarvesh Nikumbh"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: true
    anchor_sections: true
    number_tables: true
  editor_options:
    chunk_output_type: console
params:
  use_title: "archR result analysis report for CTSS in Human (inr KO 2_2)"
  sample_name: "human_cellGroup_merged"
  iqw_order_by_median: true
  data_path_on_disk: ""
  result_path_on_disk: ""
  result_dir_name: ""
  do_plot_arch: true
  do_iqw_tpm_plots: true
  do_bedfile_write: true
title: "`r params$use_title`"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE)

data_path <- file.path("data")
results_path <- file.path("results")

## Setup paths on disk
data_path_on_disk <- data_path
result_path_on_disk <- results_path
##
archR_org_data_path <- file.path(data_path, "human")
use_minTPM <- 1
archR_org_results_path <- file.path(results_path, "human")


##
sample_name <- params$sample_name
result_dir_path <- file.path(params$result_path_on_disk, params$result_dir_name)
reordered <- params$reordered
dist_method <- params$dist_method
linkage <- params$linkage
topN_val <- params$topN_val
regularize_choice <- params$regularize
iteration <- params$archR_iteration
iqw_order_by_median <- params$iqw_order_by_median



#####
do_plot_seq_image <- params$do_plot_seq_image
do_plot_arch <- params$do_plot_arch
do_IQW_TPM_plots <- params$do_iqw_tpm_plots
do_bedfile_write <- params$do_bedfile_write

```

This CAGE data was obtained from ENCODE and processed by merging all cell lines data

archR result for sample "`r sample_name`" is available at location: "<span style="color:orange">`r params$result_dir_name`</span>".


# Sample: `r sample_name` {#sample-name-section}

This file produces additional analyses plots for sample `r sample_name`, 
iteration `r iteration` of archR result for Inr knock out [-2,+2] bp.

## Parameter settings for this report

```{r param-settings}

params_df <- data.frame(param = names(params),
                        value = as.vector(unlist(params)))

knitr::kable(params_df)

```


```{r library, echo=FALSE, include=FALSE, message=FALSE}
seed_val <- 11992288
set.seed(seed_val)


library(CAGEr)
library(readr)
library(reshape2)
library(ggplot2)
library(ggeasy)
# library(forcats)
library(patchwork)
library(cowplot)
## ^ for plotting

library(ChIPseeker)
library(BSgenome.Hsapiens.UCSC.hg19)
# Use this to create one
library(GenomicFeatures)
TxDb.Hsapiens.UCSC.hg19.ensGene <- GenomicFeatures::makeTxDbFromUCSC(
                                    genome = "hg19", tablename = "ensGene")
# library(TxDb.Hsapiens.UCSC.hg19.knownGene) 
## ensGene is to be re-installed for R-4.0.0, but BiocManager says that it 
## is not available for this version of R
library(org.Hs.eg.db)
library(dendextend)
```





## Plots for sections marked TRUE are included in this report.

```{r all_funcs, include=FALSE}

source('helper-funcs.R')

```


```{r setup-variables}

sample_names <- c("human_cellGroup_merged")

archR_human_best_run <- vector("list", length(sample_names))
names(archR_human_best_run) <- sample_names

human_result <- vector("list", length(sample_names))
names(human_result) <- sample_names

human_bed_info <- vector("list", length(sample_names))
names(human_bed_info) <- sample_names

seqs_clusters_as_list <- vector("list", length(sample_names))
names(seqs_clusters_as_list) <- sample_names

seqs_clusters_as_list_ordered <- vector("list", length(sample_names))
names(seqs_clusters_as_list_ordered) <- sample_names

perSample_CAGEobj <- vector("list", length(sample_names))
names(perSample_CAGEobj) <- sample_names

perSample_peakAnno <- vector("list", length(sample_names))
names(perSample_peakAnno) <- sample_names

perSample_tissueSpec_pl <- vector("list", length(sample_names))
names(perSample_tissueSpec_pl) <- sample_names

samarth_df <- vector("list", length(sample_names))
names(samarth_df) <- sample_names
##


##
## lists storing plots
## 
## store iqw+tpm plots
perSample_pl <- vector("list", length(sample_names)) 
names(perSample_pl) <- sample_names


## store result_directory path
result_dir_path <- vector("list", length(sample_names)) 
names(result_dir_path) <- sample_names



## store clusters
perSample_archR_clusts <- vector("list", length(sample_names))
names(perSample_archR_clusts) <- sample_names


################################################################################
## Per section some variables need to be set. These are done here


# read TagClusters information from corresponding RDS file for sample
# All samples in one RDS object
cager_obj <- file.path(archR_org_data_path, 
                       paste0("samarth_hsapiens_TC_sample_human_cellGroup_merged_",
                       "minTPM", use_minTPM,
                       "_flank_up500_flank_down500_all70.rds"))

ok_chr_names <- paste0("chr", c(1:21, "X", "Y"))

for(sn in sample_names){
    result_dir_path[[sn]] <- file.path(archR_org_results_path, 
                                       paste0(sn, "_inrKO_2_2_results"))
}


```

```{r read-result-and-TCbed}



for(sn in sample_names){
    archR_human_best_run[[sn]] <- 
    file.path(archR_org_results_path, 
        paste0("archR_result_hsapiens_", sn,
               "_flank_up50_flank_down150__inrKO_2_2",
               "_modSelType_stability_chunkSize_5000", 
               "_bound_1e-06_collate_FTTTF"))
    if(file.exists(archR_human_best_run[[sn]])){
    human_result[[sn]] <- readRDS(file.path(archR_human_best_run[[sn]],
                                          "archRresult.rds"))
    }else{
    stop("Check if file exists. ", archR_human_best_run[[sn]])
    }
    ## Bed Info -- IQW and dominant TPM values 
    bed_fname <- file.path(archR_org_data_path, paste0("samarth_hsapiens_TC_sample_", sn, 
                                                   paste0("_minTPM", use_minTPM, "_all70"),
                                                   ".bed"))
    print(bed_fname)
    
    human_bed_info[[sn]] <- read.delim(file = bed_fname,
                          sep = "\t", header = TRUE)
    colnames(human_bed_info[[sn]]) <- c("chr", "start", "end", 
                                                    "IQW", "domTPM", "strand")
    ##
    message("Size of bed_info: ", nrow(human_bed_info[[sn]]))
}

```

For annotation, TSS region considered is -500 upstream and 100 nt downstream.

```{r make-cager-obj, echo=FALSE, cache=TRUE, collapse=TRUE}

# read TagClusters information from corresponding RDS file for sample
# All samples in one object

myCAGEobject <- readRDS(cager_obj)

## Per sample TCs
for(sn in sample_names){
      message("SAMPLE: ", sn)
      
      stopifnot(check_and_create_dir(result_dir_path[[sn]]))
      gr_CAGEobj_perSample_fname <- file.path(result_dir_path[[sn]], 
                                    paste("cagerObj", sn, "gr.rds", sep="_"))
      cage_sn <- paste0(sn)
      if(!file.exists(gr_CAGEobj_perSample_fname)){
          cli::cli_alert_info("Preparing new GRanges obj from CAGEr object")
      
      gr_myCAGEobject_thisSample <- GRanges(myCAGEobject)
  
      seqlengths(gr_myCAGEobject_thisSample) <- seqlengths(BSgenome.Hsapiens.UCSC.hg19)[names(seqlengths(gr_myCAGEobject_thisSample))]
      
      gr_myCAGEobject_thisSample <- trim(gr_myCAGEobject_thisSample)
      
      saveRDS(file = gr_CAGEobj_perSample_fname, 
                object = gr_myCAGEobject_thisSample)
    }else{
        cli::cli_alert_info("Using saved GRanges obj from CAGEr object")
        gr_myCAGEobject_thisSample <- readRDS(gr_CAGEobj_perSample_fname)
    }
    perSample_CAGEobj[[sn]] <- gr_myCAGEobject_thisSample
    perSample_peakAnno[[sn]] <- ChIPseeker::annotatePeak(gr_myCAGEobject_thisSample,
                                         tssRegion=c(-500, 100),
                                         TxDb=TxDb.Hsapiens.UCSC.hg19.ensGene,
                                         annoDb="org.Hs.eg.db")

}


```

## Order archR clusters by median IQ width {#prepare-objects}

By default, the clusters are ordered by their _median_ interquantile widths. 
One can choose to order by the _mean_ interquantile widths by setting paramater _iqw_order_by_median_ to `r FALSE`.


### ENCODE CAGE in Hsapiens cellGroup merged

```{r prepare-clusters-0}


## Sample 1, human_cellGroup_merged
sn <- 1
itr <- 5
use_aggl <- 'ward.D'
use_dist <- 'cor'

iter5_clusts_reord <- archR::collate_archR_result(
  result = human_result[[sn]],
  iter = itr, clust_method = 'hc', aggl_method = use_aggl,
  dist_method = use_dist, regularize = TRUE, topn = 50,
  flag = list(debugFlag = FALSE, verboseFlag = TRUE), collate = FALSE,
  return_order = TRUE)


## these are in default archR ordering
clust_archR_ord_list <- archR::get_seqs_clust_list(
  seqs_clust_lab = human_result[[ sn ]]$seqsClustLabels[[itr]])

## these are now ordered by the hc ordering
clust_hc_ord_list <- lapply(iter5_clusts_reord$order, function(x){
  clust_archR_ord_list[[x]]
})


ordered_arch_pl <- archR::plot_arch_for_clusters(
  seqs = human_result[[ sn ]]$rawSeqs,
  clust_list = clust_hc_ord_list, pos_lab = -50:150,
  xt_freq = 5, set_titles = FALSE, show = FALSE, bits_yax = "auto")

ordered_arch_pl2 <- lapply(rev(ordered_arch_pl), function(pl){
  pl <- pl +
        ggplot2::theme(axis.text = ggplot2::element_text(size = 0),
                       axis.text.x = ggplot2::element_text(
                         angle = 0, vjust = 2, hjust = 0.5),
                       axis.text.y = ggplot2::element_text(vjust = 0.5),
                       axis.title.y = ggplot2::element_text(size = 0),
                       axis.ticks.length = ggplot2::unit(0.00, "cm"),
                       plot.margin = ggplot2::unit(c(-0.1,0,-0.4,-0.4), "cm"))
})
sam_foo <- cowplot::plot_grid(plotlist = ordered_arch_pl2, ncol = 1)

use_cutk <- 20

stopifnot(check_and_create_dir(result_dir_path[[sn]]))
fname <- file.path(result_dir_path[[sn]], paste0(sample_names[sn], "_dend_arch_list_",
                                           use_dist, "_", use_aggl, "_",
                                           use_cutk, "clusters"))

sam_foo2 <- plot_dend_arch(arch_plot = sam_foo, fname = fname, use_cutk = use_cutk,
              clusts = iter5_clusts_reord, use_ht = 60, plot_png = FALSE,
              lwd = 0.4, repel = TRUE, show_labels = TRUE, labels_track_height = 0.25,
              rect = TRUE, rect_fill = TRUE,
              color_labels_by_k = TRUE)

## IMPORTANT: Seems like no additional manipulation is required. 


# 
# 
temp_clusts <- cutree(iter5_clusts_reord, k = use_cutk)
names(temp_clusts) <- NULL
## Make further few clusters
# nCl <- length(unique(temp_clusts))
# 
# # TATA architectures
# temp_clusts[c(34)] <- nCl + 1
# temp_clusts[c(33)] <- nCl + 2
# temp_clusts[c(35)] <- nCl + 3
# temp_clusts[c(26)] <- nCl + 4
# 
# # TCT architecture
# temp_clusts[c(37)] <- nCl + 5
# 
# temp_clusts[c(31)] <- nCl + 6
# temp_clusts[c(36)] <- nCl + 6
# 
# temp_clusts[c(17)] <- temp_clusts[c(13)]

# Re-plot with proper coloring that shows the manipulations in clusters
use_cutk <- 20

stopifnot(check_and_create_dir(result_dir_path[[sn]]))
fname2 <- file.path(result_dir_path[[sn]], paste0(sample_names[sn], "_dend_arch_list_",
                                           use_dist, "_", use_aggl, "_",
                                           use_cutk, "clusters_final"))


## Also need to show alongside, how the final clusters' seqlogos look
clust_list <- lapply(unique(temp_clusts), function(x){which(temp_clusts == x)})
seqs_clusters_as_list[[sample_names[sn]]] <- archR::collate_clusters(clust_list, archR::get_seqs_clust_list(human_result[[sn]]$seqsClustLabels[[itr]]))
##

use_color <- scales::hue_pal()(length(unique(temp_clusts)))
sam_foo2 <- plot_dend_arch(arch_plot = sam_foo, fname = fname2, use_ht = 60,
               use_cutk = use_cutk,#length(unique(temp_clusts)), 
               clusts = iter5_clusts_reord, rect = TRUE, rect_fill = TRUE,
               label_cols = use_color[temp_clusts[iter5_clusts_reord$order]],
               k_colors = use_color, 
               clust_assignment = clust_list,
               new_clusts = seqs_clusters_as_list[[sample_names[sn]]],
               rawSeqs = human_result[[sample_names[sn]]]$rawSeqs,
               palette = FALSE, plot_png = FALSE)


cluster_medians_IQW <- unlist(lapply(seqs_clusters_as_list[[sn]], function(x){
    median(human_bed_info[[sn]]$IQW[x])
}))
cluster_means_IQW <- unlist(lapply(seqs_clusters_as_list[[sn]], function(x){
    mean(human_bed_info[[sn]]$IQW[x])
}))

if(iqw_order_by_median){
    ascending_order_IQW <- sort(cluster_medians_IQW, decreasing = FALSE,
                            index.return = TRUE)
}else{
    ascending_order_IQW <- sort(cluster_means_IQW, decreasing = FALSE,
                            index.return = TRUE)
}
##
seqs_clusters_as_list_ordered[[sn]] <-
                      lapply(ascending_order_IQW$ix,
                              function(x){
                                  seqs_clusters_as_list[[sn]][[x]]
                              })

perSample_archR_clusts[[sn]] <- seqs_clusters_as_list_ordered[[sn]]


```



```{r make-df-from-bed}

## Using Tau as a measure of tissue specificity
## This measure is defined/used in this paper:
##  https://www.aging-us.com/article/202648/text
## Various tissue specificity measures have been reviewed in this paper:
## https://doi.org/10.1093/bib/bbw008
## The above benchmarking paper finds Tau to be the most robust and sensitive 
## measure of tissue specificity. 
## 
tissueSpecTau <- read_csv(file.path(data_path, "tissueSpecificity-tau", 
                                    "Tau_score", "Tau_gene_V8.csv"), 
                          col_names = TRUE, show_col_types = FALSE)


tau_dist <- ggpubr::gghistogram(tissueSpecTau, x = "tau", 
    title = "Tau distribution/all genes (ensembl ids); N = 56156")

ggplot2::ggsave(file.path(result_dir_path[[sn]], "tau_distribution.pdf"), 
    plot = tau_dist, device = "pdf", width = 6, height = 4)

for(sn in sample_names) {
    clust_lens <- unlist(lapply(seqs_clusters_as_list_ordered[[sn]], length))
    message("Ordered list lengths: ", paste(clust_lens, collapse = " "))
    message("Original list lengths:", 
      paste(unlist(lapply(seqs_clusters_as_list[[sn]], length)), collapse = " "))
  
  
    ##
    clust_lab <- rep("0", length(human_result[[sn]]$rawSeqs))
    clust_names <- sort(as.character(1:length(seqs_clusters_as_list_ordered[[sn]])))
    for(i in seq_along(seqs_clusters_as_list_ordered[[sn]])){
        clust_lab[seqs_clusters_as_list_ordered[[sn]][[i]] ] <- clust_names[i]
    }
  
    peakAnno_df <- as.data.frame(perSample_peakAnno[[sn]])
    
    
    
    tissSpecScores <- vapply(seq(nrow(peakAnno_df)), FUN = function(x){
        idx <- which(tissueSpecTau$gene_id == peakAnno_df[x, "geneId"])
        if(length(idx) < 1){
            return(-1)
        }
        return(tissueSpecTau$tau[idx])
        }, FUN.VALUE=numeric(1))

    message("Tissue specificity values (tau) collected")
    
    
    ### Get DF ob BED information
    samarth_df[[sn]] <- data.frame(chr = human_bed_info[[sn]]$chr, 
                             start = human_bed_info[[sn]]$start,
                             end = human_bed_info[[sn]]$end,
                             strand = human_bed_info[[sn]]$strand,
                             IQW = human_bed_info[[sn]]$IQW,
                             domTPM = human_bed_info[[sn]]$domTPM,
                             
                             geneId = peakAnno_df$geneId,
                             tissueSpecScore = tissSpecScores,
                             clust_ID = clust_lab #these are character sorted
                             )
    
    df_fname <- file.path(result_dir_path[[sn]], "samarth_info_df.csv")
    write.csv(samarth_df, file = df_fname)
    
    ## All, includes repeated ones
    fname <- file.path(result_dir_path[[sn]], "tau_existing_all_genes_distribution.pdf")
    p1 <- ggpubr::gghistogram(data.frame(tissSpecScores = tissSpecScores), 
                        x = "tissSpecScores",  y = "..count..")
    ggplot2::ggsave(filename = fname, plot = p1, device = pdf, width = 6, height = 4)
    
    ## Just for the 4th cluster which looks like TATA-box
    fname <- file.path(result_dir_path[[sn]], "tau_clust_4_distribution.pdf")
    p1 <- ggpubr::gghistogram(
        data.frame(tissSpecScores_clust4 = tissSpecScores[perSample_archR_clusts[[1]][[4]]]), 
        x = "tissSpecScores_clust4", y = "..count..")
    ggplot2::ggsave(filename = fname, plot = p1, device = pdf, width = 6, height = 4)
    
    
}


```



# Session Info

```{r sessionInfo, echo=FALSE}

sessionInfo()

```
